---
title: "E1: Correctly Specified DGP"
format:
  html:
    mainfont: "Times New Roman"
editor_options: 
  chunk_output_type: inline
---

```{r}
#| message: false
library(tidyverse)
library(kableExtra)
library(gt)


df <- bind_rows(
  read.csv("E1_results.csv.gz"),
  read.csv("Mispeccified_results.csv.gz"),
  read.csv("E2_results.csv.gz"),
)
```

# Expected Results

**E1 (Correct specification, strong overlap):** Both methods should produce similar bias and RMSE. Any divergence here signals an implementation or DGP issue rather than a substantive finding. This serves as a sanity check for the simulation infrastructure [@keele2025; @tan2020b].

**E2 (Overlap degradation):** As the overlap scaler $c$ decreases, bias and RMSE for `cv.glmnet` should increase faster than for `cv.balnet`. The performance gap should widen as overlap weakens [@keele2025; @chattopadhyay2020].

**E3 (Misspecification):** Under misspecification via Kang--Schafer transformations, `cv.balnet` should show a clear advantage in both bias and RMSE, exceeding the gap observed under correct specification. The interaction of misspecification with weak overlap should produce the largest relative advantage. This follows from the calibration loss controlling MSRE, which bounds IPW bias via Cauchy--Schwarz, whereas the MLE loss controls KL divergence, which does not directly bound IPW bias [@tan2020b, Prop. 3].

**E4 (High dimensionality):** With $s$ fixed, increasing $p$ should hurt both methods, but `cv.balnet` should be more robust. Under misspecification at high $p$, RCAL selects sparser models than RML while achieving equal or better IPW estimation [@tan2020b].

**E5 (Soft sparsity):** As $s$ increases with harmonic decay coefficients, the boundary between active and inactive covariates blurs under the AR(1) covariance structure. The balancing loss should handle this more robustly because it optimises for balance rather than prediction, retaining weakly-correlated covariates that happen to improve balance. There is no direct precedent in the project literature for this experiment; it is the simulation's novel contribution.

**E6 (Outcome complexity):** More complex outcome functions (quadratic, exponential) should increase RMSE for both methods, but the relative advantage of the balancing approach should persist, especially under misspecification [@tan2020b].

**E7 (Lambda interpretation and fixed-lambda behaviour):** Fixed $\lambda = 0.05$ should produce tighter balance but higher variance (lower effective sample size) than $\lambda = 0.10$. The CV-selected $\lambda$ should generally fall between these. Under weak overlap, fixed $\lambda = 0.05$ may become infeasible due to path truncation, which is itself a diagnostic for positivity violations [@sverdrup2026].

**E8 (Computational cost):** `cv.balnet` must fit two separate propensity score models for ATE estimation (one for each potential outcome mean), roughly doubling computation relative to `cv.glmnet`'s single model [@sverdrup2026, Remark 1].

**E9 (Variable selection):** `cv.balnet` should consistently select fewer nonzero coefficients than `cv.glmnet`, especially under misspecification. This is consistent with the calibration loss requiring fewer covariates to achieve balance than the MLE loss requires for predictive accuracy [@tan2020b].

# 

# DGP 1

### Data-Generating Process

We draw $n$ observations of $p$ covariates $X_i = (X_{i1}, \dots, X_{ip})^\top \sim \mathcal{N}(0, \Sigma)$, where $\Sigma_{jk} = \rho^{|j-k|}$ with $\rho = 0$ (independent) or $\rho = 0.5$ (AR(1)). The true propensity score is

$$e^*(X_i) = \left\{1 + \exp\left(\frac{X_{i1} - 0.5\,X_{i2} + 0.25\,X_{i3} + 0.1\,X_{i4}}{c}\right)\right\}^{-1}$$

where $c > 0$ is an overlap scaler [@keele2025], and treatment is assigned as $W_i \sim \text{Bernoulli}(e^*(X_i))$. Only the first four of $p$ covariates are active in the propensity score, creating a sparse signal setting consistent with @tan2020.

Under the sharp null $\tau = 0$, potential outcomes are generated as

$$\begin{aligned}
Y_i^{\text{lin}} &= X_{i1} + 0.5\,X_{i2} + 0.5\,X_{i3} + 0.5\,X_{i4} + \varepsilon_i \\
Y_i^{\text{quad}} &= \sum_{j=1}^{4} \{\max(X_{ij}, 0)\}^2 + \varepsilon_i \\
Y_i^{\text{exp}} &= \sum_{j=1}^{4} \exp(X_{ij}/2) + \varepsilon_i
\end{aligned}$$

where $\varepsilon_i \sim \mathcal{N}(0, 1)$. The linear specification follows @tan2020 (lin1); the quadratic and exponential forms are also drawn from @tan2020.

To introduce misspecification, the analyst observes transformed covariates $\tilde{X}$ instead of $X$ [@kangschafer2007]:

$$\begin{aligned}
\tilde{X}_{i1} &= \exp(X_{i1}/2) \\
\tilde{X}_{i2} &= \frac{X_{i2}}{1 + \exp(X_{i1})} + 10 \\
\tilde{X}_{i3} &= (0.04\,X_{i1}\,X_{i3} + 0.6)^3 \\
\tilde{X}_{i4} &= (X_{i2} + X_{i4} + 20)^2
\end{aligned}$$

where $\tilde{X}_{i1}, \dots, \tilde{X}_{i4}$ are subsequently standardized to have mean zero and unit variance, following @tan2020. For $j = 5, \dots, p$, $\tilde{X}_{ij} = X_{ij}$. Both logistic and linear regressions on $\tilde{X}$ appear adequate by standard diagnostics, yet are misspecified since the true models are linear in $X$ on the logit scale.

The simulation varies sample size $n \in \{500, 1000, 5000\}$, covariate dimension $p \in \{10, 50, 100, \dots\}$, and overlap $c \in \{2, 1, 0.75, 0.5, 0.25\}$, with each experiment isolating one factor at a time.

## Replicate Tan in RCAL (same DGP)

**Tan's core result replicates.** Under misspecification, balnet outperforms glmnet across all three outcome functions, with the advantage growing with n and largest for quad1 (5.6×) and exp (2.6×). RML-HT deteriorates with n while RCAL stays flat. This validates simulation infrastructure and the calibration loss advantage.

| $n$ | Tan RCAL $\times \sqrt{2}$ | bal0  | bal05 | balnetcv |
|-----|----------------------------|-------|-------|----------|
| 200 | 0.198                      | 0.173 | 0.206 | 0.231    |
| 400 | 0.127                      | 0.124 | 0.173 | 0.192    |
| 800 | 0.085                      | 0.089 | 0.156 | 0.176    |

: Convergence comparison: RMSE under correct specification ($p=4$, linear outcome, $c=1$, iid covariates, 1000 replications). Tan RCAL values from Tan (2020, Table S3) scaled by $\sqrt{2}$ for the ATE estimand.\
\
**bal0 recovers Tan's convergence rate exactly.** After √2 scaling for the ATE vs μ¹ estimand difference, bal0 RMSE at (200/400/800, p=4, correct spec, linear) is 0.173/0.124/0.089 versus the expected 0.198/0.127/0.085.

- **bal0 converges with Tan**; bal05 and balnetcv flatten out.
  - This isolates the attenuation to cv.balnet's λ selection and confirms the balnet loss function itself is correct.
- Tan's CV behaves somewhere between bal0 and bal05 depending on the outcome, but consistently below balcv's selected λ.
- From Tan (2020, p.19), his CV minimises the held-out calibration loss over a discrete grid. **change CV criterion?**

## Setup

- **Grid:** n ∈ {500, 1000, 5000, 10000, 50000}, p = 50, overlap = 1
- **Outcomes:** linear, quad1, exp
- **Misspec:** FALSE
- **Estimators:** balnet (cv, λ→0, λ=0.05, λ=0.10), glmnet (HT, Hajek), oracle (HT, Hajek)
- **Reps:** 1000

## Summary

```{r}
summary_df <- df |>
  group_by(n, p, misspec, outcome, overlap, estimator) |>
  summarise(
    m          = n(),
    bias       = mean(tau_hat),
    mcse_bias  = sd(tau_hat) / sqrt(m),
    empse      = sd(tau_hat),
    mcse_empse = empse / sqrt(2 * (m - 1)),
    rmse       = sqrt(mean(tau_hat^2)),
    mcse_rmse  = sd(tau_hat^2) / (2 * rmse * sqrt(m)),
    n_extreme  = sum(abs(tau_hat) > 10),
    .groups    = "drop"
  )
```

## E1: Correct Specification

```{r}

label_map <- c(
  "oracle"            = "Oracle",
  "normalised_oracle" = "Normalised Oracle",
  "balnet0"           = "BalNet (λ=0)",
  "balnet05"          = "BalNet (λ=.05)",
  "balnet10"          = "BalNet (λ=.10)",
  "balnetcv"          = "BalNet (CV)",
  "glmnetcv"          = "GLMNet (CV)",
  "glmcv_normal"      = "GLM (CV, Normalized)"
)

e1_estimators <- c("balnetcv", "balnet05", "balnet0", "glmnetcv", "normalised_oracle")

estimator_colours <- c(
  "BalNet (CV)"        = "#2ca02c",  # green
  "BalNet (λ=.05)"     = "#1f77b4",  # blue
  "BalNet (λ=0)"       = "#17becf",  # teal
  "GLMNet (CV)"        = "#d62728",  # red
  "GLM (CV, Normalized)" = "#d62728",  # red
  "Normalised Oracle"  = "#7f7f7f"   # grey
)

# Re-plot E1 (overlap = 1 only)
plot_rmse <- function(data, outcome_name, estimators, label_map, colours,
                      title = NULL, is_misspec = FALSE) {
  data |>
    dplyr::filter(misspec == is_misspec, outcome == outcome_name, 
                  estimator %in% estimators, overlap == 1) |>
    dplyr::mutate(estimator = forcats::fct_relevel(
      dplyr::recode(estimator, !!!label_map), label_map[estimators])) |>
    ggplot2::ggplot(ggplot2::aes(n, rmse, colour = estimator)) +
    ggplot2::geom_line() +
    ggplot2::geom_point(size = 2) +
    ggplot2::scale_x_log10(breaks = c(500, 1000, 5000, 10000, 50000),
                           labels = scales::comma) +
    ggplot2::scale_y_continuous(limits = c(0, NA)) +
    ggplot2::scale_colour_manual(values = colours) +
    ggplot2::labs(x = "n", y = "RMSE", colour = "Estimator",
                  title = title %||% paste0(stringr::str_to_title(outcome_name), " — RMSE")) +
    ggplot2::theme_minimal()
}

plot_rmse(summary_df, "linear", e1_estimators, label_map, estimator_colours,
          title = "E1: Correct Specification (p = 50, c = 1)")

#table

make_sim_table <- function(data, outcome_name, estimators, label_map,
                           title = NULL, is_misspec = FALSE, overlap_val = 1) {
  tbl <- data |>
    filter(misspec == is_misspec, outcome == outcome_name,
           estimator %in% estimators, overlap == overlap_val) |>
    mutate(
      estimator = factor(estimator, levels = names(label_map), labels = label_map),
      cell = paste0(sprintf("%.3f", bias), "<br><small>(", sprintf("%.3f", rmse), ")</small>"),
      avg_rmse = ave(rmse, estimator, FUN = mean)
    )
  
  tbl |>
    dplyr::select(estimator, n, cell, avg_rmse) |>
    pivot_wider(names_from = n, values_from = cell, names_prefix = "n = ") |>
    arrange(avg_rmse) |>
    dplyr::select(-avg_rmse) |>
    kbl(caption = title %||% paste0(str_to_title(outcome_name), " — Bias (RMSE)"), escape = FALSE) |>
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE, latex_options = c("hold_position", "scale_down")
    ) |>
    column_spec(1, bold = TRUE) |>
    footnote(general = "Top: Bias. Bottom (smaller): RMSE. Ordered by average RMSE.",
             general_title = "")
}

make_sim_table(summary_df, "linear", e1_estimators, label_map,
               title = "E1: Correct Specification, Linear — Bias (RMSE)")
```

**E1 holds at the loss-function level, bal0 matches or beats glmnet at every n, but cv.balnet selects λ large enough to erase the advantage.**

- **bal0 more efficient than normalized oracle** — consistent with Wager (2024). Confirms the calibration loss function is correctly implemented and dominates MLE at every n.

-  **cv.balnet over-regularizes, erasing the loss function advantage.** CV selects λ too large → non-vanishing attenuation bias ≈ −0.13. This is the mechanism worth investigating — not a bug, but a cost of the regularization that pays off under misspecification/overlap degradation (E2, E3).

- **bal05 outperforms glmnet at n ≤ 1000, glmnet overtakes by n = 5,000.** Expected, bal05's fixed bias floor (−0.14) is cheaper than glmnet's variance at small n, but glmnet is consistent so it eventually wins. correct specification rewards consistency.

- **Outcome complexity affects attenuation magnitude, not ranking.** Bias floors: linear ≈ −0.14, quad1 ≈ −0.15, exp ≈ −0.11. Same story across all three.

- **check?** CV-selected λ at each n. You noted CV selects λ ≈ 0.23 at n=5,000 for p=10 — reporting the p=50 equivalent would quantify how much CV over-regularises and connect directly to the KKT interpretation (λ = max allowable imbalance).

## With Overlap

```{r}
library(ggh4x)

plot_data_final <- df |>
  filter(
    outcome == "linear",
    n %in% c(500, 5000, 10000),
    overlap %in% c(1, 0.5, 0.25),
    estimator %in% c("balnet05", "glmcv_normal")
  ) |>
  mutate(
    estimator = dplyr::recode(estimator,
                              "balnet05" = "BalNet (\u03bb = 0.05)",
                              "glmcv_normal" = "GLM (CV, Normalized)"),
    n_label = factor(paste0("n = ", scales::comma(n)),
                     levels = paste0("n = ", scales::comma(c(500, 5000, 10000)))),
    overlap = factor(paste0("c = ", overlap),
                     levels = paste0("c = ", c(1, 0.5, 0.25)))
  )

ggplot(plot_data_final, aes(x = tau_hat, fill = estimator, colour = estimator)) +
  geom_density(alpha = 0.35, linewidth = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4) +
  facet_grid2(overlap ~ n_label, scales = "free", independent = "all") +
  labs(
    title = "Linear Outcome",
    x = "Estimate", y = "Density", fill = "Estimator", colour = "Estimator"
  ) +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 3)) +
  scale_y_continuous(breaks = scales::breaks_pretty(n = 3)) +
  scale_fill_manual(values = c("BalNet (\u03bb = 0.05)" = "#E69F00",
                               "GLM (CV, Normalized)" = "#56B4E9")) +
  scale_colour_manual(values = c("BalNet (\u03bb = 0.05)" = "#E69F00",
                                 "GLM (CV, Normalized)" = "#56B4E9")) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 11),
    strip.text.x = element_text(face = "bold", size = 12),
    strip.text.y = element_text(face = "bold", size = 11, angle = 0),
    strip.clip = "off",
    panel.grid.minor = element_blank(),
    panel.spacing = unit(0.8, "lines"),
    plot.margin = margin(5, 20, 5, 5)
  )

plot_data_balnet <- df |>
  filter(
    outcome == "linear",
    n %in% c(500, 5000, 10000),
    overlap %in% c(1, 0.5, 0.25),
    estimator %in% c("balnet05", "balnetcv")
  ) |>
  mutate(
    estimator = dplyr::recode(estimator,
                              "balnet05" = "BalNet (\u03bb = 0.05)",
                              "balnetcv" = "BalNet (CV)"),
    n_label = factor(paste0("n = ", scales::comma(n)),
                     levels = paste0("n = ", scales::comma(c(500, 5000, 10000)))),
    overlap = factor(paste0("c = ", overlap),
                     levels = paste0("c = ", c(1, 0.5, 0.25)))
  )

ggplot(plot_data_balnet, aes(x = tau_hat, fill = estimator, colour = estimator)) +
  geom_density(alpha = 0.35, linewidth = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4) +
  facet_grid2(overlap ~ n_label, scales = "free", independent = "all") +
  labs(
    title = "Balnet .05 vs cvbalnet",
    x = "Estimate", y = "Density", fill = "Estimator", colour = "Estimator"
  ) +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 3)) +
  scale_y_continuous(breaks = scales::breaks_pretty(n = 3)) +
  scale_fill_manual(values = c("BalNet (\u03bb = 0.05)" = "#E69F00",
                               "BalNet (CV)" = "#56B4E9")) +
  scale_colour_manual(values = c("BalNet (\u03bb = 0.05)" = "#E69F00",
                                 "BalNet (CV)" = "#56B4E9")) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 11),
    strip.text.x = element_text(face = "bold", size = 12),
    strip.text.y = element_text(face = "bold", size = 11, angle = 0),
    strip.clip = "off",
    panel.grid.minor = element_blank(),
    panel.spacing = unit(0.8, "lines"),
    plot.margin = margin(5, 20, 5, 5)
  )

plot_data_bal0v05 <- df |>
  filter(
    outcome == "linear",
    n %in% c(500, 5000, 10000),
    overlap %in% c(1, 0.5, 0.25),
    estimator %in% c("balnet0", "balnet05")
  ) |>
  mutate(
    estimator = dplyr::recode(estimator,
                              "balnet0" = "BalNet (\u03bb = 0)",
                              "balnet05" = "BalNet (\u03bb = 0.05)"),
    n_label = factor(paste0("n = ", scales::comma(n)),
                     levels = paste0("n = ", scales::comma(c(500, 5000, 10000)))),
    overlap = factor(paste0("c = ", overlap),
                     levels = paste0("c = ", c(1, 0.5, 0.25)))
  )

ggplot(plot_data_bal0v05, aes(x = tau_hat, fill = estimator, colour = estimator)) +
  geom_density(alpha = 0.35, linewidth = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4) +
  facet_grid2(overlap ~ n_label, scales = "free", independent = "all") +
  labs(
    title = "0.5 vs 0 BalNet",
    x = "Estimate", y = "Density", fill = "Estimator", colour = "Estimator"
  ) +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 3)) +
  scale_y_continuous(breaks = scales::breaks_pretty(n = 3)) +
  scale_fill_manual(values = c("BalNet (\u03bb = 0)" = "#E69F00",
                               "BalNet (\u03bb = 0.05)" = "#56B4E9")) +
  scale_colour_manual(values = c("BalNet (\u03bb = 0)" = "#E69F00",
                                 "BalNet (\u03bb = 0.05)" = "#56B4E9")) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 11),
    strip.text.x = element_text(face = "bold", size = 12),
    strip.text.y = element_text(face = "bold", size = 11, angle = 0),
    strip.clip = "off",
    panel.grid.minor = element_blank(),
    panel.spacing = unit(0.8, "lines"),
    plot.margin = margin(5, 20, 5, 5)
  )
```

\

- **bal05 outperforms glmnet at every sample size when overlap is weak (c = 0.25)**, with the advantage growing with n. At moderate overlap (c = 0.5), glmnet's consistency advantage eventually overtakes bal05 at large n for the linear outcome, but not for the nonlinear outcomes.

- **at overlap c=0.25, glmnet RMSE increases with N.**

- **Under correct specification, bal0 outpeforms bal05 and everything else,** at every overlap level. loss function handles weak overlap cleanly.

- **cv.balnet:** Despite over-regularizing, cv.balnet still outperforms glmnet-Hajek in the majority of weak-overlap cells.

- **Outcome complexity:** The pattern holds across all three outcome functions. Outcome complexity amplifies glmnet's divergence under weak overlap.

- **balnet predictable bias:** balnets attenuation bias is stable and predictable, robust estimation through AIPW should help account for this?

## With Misspeccification

```{r}
library(tidyverse)
library(kableExtra)

raw_ms  <- read.csv("Mispeccified_results.csv.gz")
raw_ms2 <- read.csv("Mispeccified_Lineartest_results2.csv.gz")

ests_ms <- c("balnet0","balnet05","balnet10","balnetcv","glmcv_normal","glmnetcv")
e2_ests <- c("balnetcv","balnet05","balnet0","glmcv_normal")
e3_ests <- c("balnetcv","balnet05","balnet0","balnet10","glmcv_normal","glmnetcv")

label_map_ms <- c(balnet0="BalNet (\u03bb=0)", balnet05="BalNet (\u03bb=.05)",
                  balnet10="BalNet (\u03bb=.10)", balnetcv="BalNet (CV)",
                  glmcv_normal="GLM (CV, Normalized)", glmnetcv="GLMNet (CV)")
color_map_ms <- c("BalNet (CV)"="#2ca02c", "BalNet (\u03bb=.05)"="#1f77b4",
                  "BalNet (\u03bb=0)"="#17becf", "BalNet (\u03bb=.10)"="#ff7f0e",
                  "GLM (CV, Normalized)"="#d62728", "GLMNet (CV)"="#9467bd")

make_summ <- function(data) {
  data |>
    filter(estimator %in% ests_ms) |>
    group_by(n, p, misspec, outcome, overlap, estimator) |>
    summarise(m = n(), bias = mean(tau_hat), mcse_bias = sd(tau_hat)/sqrt(m),
              empse = sd(tau_hat), mcse_empse = empse/sqrt(2*(m-1)),
              rmse = sqrt(mean(tau_hat^2)), mcse_rmse = sd(tau_hat^2)/(2*rmse*sqrt(m)),
              n_extreme = sum(abs(tau_hat) > 10), .groups = "drop")
}

summ_ms  <- make_summ(raw_ms)
summ_ms2 <- make_summ(raw_ms2)

theme_ms <- theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"),
        panel.grid.minor = element_blank())

plot_rmse_ms <- function(data, outcome_name, estimators, title = NULL) {
  data |>
    filter(misspec == TRUE, outcome == outcome_name,
           estimator %in% estimators, overlap == 1) |>
    mutate(estimator = factor(estimator, names(label_map_ms), label_map_ms)) |>
    ggplot(aes(n, rmse, colour = estimator)) +
    geom_line(linewidth = 0.8) + geom_point(size = 2.5) +
    scale_x_log10(breaks = c(500,1000,5000,10000,50000), labels = scales::comma) +
    scale_y_continuous(limits = c(0, NA)) +
    scale_colour_manual(values = color_map_ms) +
    labs(x = "n", y = "RMSE", colour = "Estimator",
         title = title %||% paste0(str_to_title(outcome_name), " — RMSE")) +
    theme_ms
}

make_table_ms <- function(data, outcome_name, estimators, title = NULL) {
  tbl <- data |>
    filter(misspec == TRUE, outcome == outcome_name,
           estimator %in% estimators, overlap == 1) |>
    mutate(estimator = factor(estimator, names(label_map_ms), label_map_ms),
           cell = paste0(sprintf("%.3f", bias), "<br><small>(", sprintf("%.3f", rmse), ")</small>"),
           avg_rmse = ave(rmse, estimator, FUN = mean))
  tbl |>
    select(estimator, n, cell, avg_rmse) |>
    pivot_wider(names_from = n, values_from = cell, names_prefix = "n = ") |>
    arrange(avg_rmse) |> select(-avg_rmse) |>
    kbl(caption = title %||% paste0(str_to_title(outcome_name), " — Bias (RMSE)"),
        escape = FALSE) |>
    kable_styling(bootstrap_options = c("striped","hover","condensed"),
                  full_width = FALSE, latex_options = c("hold_position","scale_down")) |>
    column_spec(1, bold = TRUE) |>
    footnote(general = "Top: Bias. Bottom (smaller): RMSE. Ordered by average RMSE.",
             general_title = "")
}

# RMSE plots — linear from summ_ms2, rest from summ_ms
plot_rmse_ms(summ_ms, "linear",  e2_ests, "Misspecified quad (p = 50, c = 1)")
plot_rmse_ms(summ_ms2, "linear", e2_ests, "Misspecified (p = 50, c = 1)") +
  scale_x_log10(breaks = c(250,500,750,1000,2000,3000,4000,5000), labels = scales::comma)
plot_rmse_ms(summ_ms, "quad1",  e2_ests, "Misspecified quad (p = 50, c = 1)")
plot_rmse_ms(summ_ms, "exp",    e2_ests, "Misspecified exp (p = 50, c = 1)")

# Tables — linear from summ_ms2, rest from summ_ms
make_table_ms(summ_ms2, "linear", e3_ests, "Misspecified, Linear — Bias (RMSE)")
make_table_ms(summ_ms, "quad1",  e3_ests, "Misspecified, Quadratic — Bias (RMSE)")
make_table_ms(summ_ms, "exp",    e3_ests, "Misspecified, Exponential — Bias (RMSE)")


# bias plots 

plot_bias_ms(summ_ms, "linear",  e2_ests, "Misspecified linear (p = 50, c = 1) — |Bias|")
plot_bias_ms(summ_ms2, "linear", e2_ests, "Misspecified linear (p = 50, c = 1) — |Bias|") +
  scale_x_log10(breaks = c(250,500,750,1000,2000,3000,4000,5000), labels = scales::comma)
plot_bias_ms(summ_ms, "quad1",  e2_ests, "Misspecified quad (p = 50, c = 1) — |Bias|")
plot_bias_ms(summ_ms, "exp",    e2_ests, "Misspecified exp (p = 50, c = 1) — |Bias|")

# Cell wins (with ties)
summ_ms |>
  filter(misspec == TRUE, overlap == 1) |>
  group_by(outcome, n) |>
  filter(rmse == min(rmse)) |>
  ungroup() |>
  count(estimator, name = "cells_won")



#RMSE decomp
decomp_abs_all <- summ_ms |>
  filter(misspec == TRUE, overlap == 1,
         outcome %in% c("linear", "quad1", "exp"),
         estimator %in% c("balnet0", "balnet05", "balnetcv", "glmcv_normal")) |>
  mutate(
    bias_sq = bias^2,
    var     = empse^2,
    estimator = factor(estimator, names(label_map_ms), label_map_ms),
    outcome = factor(outcome, c("linear", "quad1", "exp"),
                     c("Linear", "Quadratic", "Exponential"))
  ) |>
  select(n, outcome, estimator, bias_sq, var) |>
  pivot_longer(cols = c(bias_sq, var), names_to = "component", values_to = "value") |>
  mutate(component = factor(component, c("var", "bias_sq"), c("Variance", "Bias²")))

decomp_prop_all <- decomp_abs_all |>
  group_by(n, outcome, estimator) |>
  mutate(prop = value / sum(value)) |>
  ungroup()

ggplot(decomp_prop_all, aes(x = n, y = prop, fill = component)) +
  geom_area(alpha = 0.8) +
  facet_grid(estimator ~ outcome) +
  scale_x_log10(breaks = c(500, 5000, 50000), labels = scales::comma) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Bias²" = "#e41a1c", "Variance" = "#377eb8")) +
  labs(x = "n", y = "Share of MSE", fill = NULL,
       title = "MSE Decomposition (%) — Misspecified (c = 1)") +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    strip.text.y = element_text(face = "bold", size = 9, angle = 0),
    strip.text.x = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(5, 50, 5, 5)
  )
```

**E3 (Misspecification):** Under misspecification via Kang--Schafer transformations, `cv.balnet` should show a clear advantage in both bias and RMSE, exceeding the gap observed under correct specification. The interaction of misspecification with weak overlap should produce the largest relative advantage. This follows from the calibration loss controlling MSRE, which bounds IPW bias via Cauchy--Schwarz, whereas the MLE loss controls KL divergence, which does not directly bound IPW bias [@tan2020b, Prop. 3].

As expected, under misspecification balancing methods outperformed RMSE estimates against the MLE approach consistently. However, at lower $n$ bias for non-linear outcomes either matches our outperforms balancing estimators before exploding as $n \to \infty$. This reflects the RMSE decomposition, where balancing estimators are bias driven and MLE ones are variance driven with bias increasing with sample size.

Under doubly robust estimators, it would be interesting to see how balancing methods inherent negative bias is accounted for, and if these results change.

Within the balnet package, it seems like under misspeccification alone that setting $\lambda=0$ produces the best results consistently, however cv.balnet has improved comparitivly to when the model was correctly specified.

#### !!Overall, balancing VS MLE!!

-  **bal0 RMSE outperforms every feasible estimator in all cells, with one tie:**

  -  balancing loss function advantage holds under misspecification.

    - *in E1 bal0 wins on efficiency (beats the oracle); in E3 it wins on robustness (bounded while MLE diverges, but the misspec bias floor means it no longer consistently beats the oracle\*).*

      - *\*still beats oracle under quad1*

  - *bal10 edges bal0 in the most variance dominated cell, quad1 n=500 (ahead by 0.002,
    0.5 mcse - consistent with simulation noise; resolving it would take \~45k reps). everywhere else bal0 wins outright.*

- **normalised MLE diverges with n under misspecification (K&S amplification); all balnet variants stay bounded.**

  - Generally for MLE, RMSE divergence is primarily variance driven. For nonlinear outcomes, bias also grows with n, compounding the divergence.

  - *Divergence form is outcome dependent. Bias explodes for nonlinear outcomes (+1.15 at n=50000 for quad1, +0.38 exp). For linear, divergence is variance driven, bias approaches zero around n=5000 before increasing, and remains below all balnet variants through n=50000.*

  - *In the exponential case, bias only overtakes balnet after n≈10000, while for quadratic MLE bias overtakes balnet earlier around n≈1000.*

  - non normalised MLE explodes, will be excluded from further analysis.

  - **for balancing estimators mispeccification introduced its own bias, seperate from regularization attenuation:** *bal0 bias doesn't vanish under misspec (Linear worsens, -0.18 → −0.21, quad1 improves −0.10 → −0.04, exp stable at−0.12), where for correctly specified* *bal0 → 0.*

- **MLE RMSE is primarily variance driven while balancing estimator RMSE is bias driven**

#### !!Within Balancing Estimators!!

- **For balancing estimators, larger λ means larger rmse: bal0 wins in every cell\***

  - varience reduction from increasing $\lambda$ is small, effect on RMSE is outweighed by increase in bias.

    - \**exception is with a quadratic outcome at n=500, where bal10 RMSE* $\approx$ *bal0 RMSE.*

    - *at n=500 for nonlinear outcomes, bal05 bias is* $\approx$ *bal10, but bal10 has slightly less variance from stronger regularization. So bal10 wins. bal0 has meaningfully less bias than both, so it still wins overall.*

- **the criterion cv.balnet uses (balancing loss) fails to select** $\lambda$ **competitive against optimal** $\lambda$ **path endpoint.**

  - cv.balnet's $\lambda_{cv}$ shrinks toward the path endpoint as n grows under misspec, unlike under correct spec where $\lambda_{cv}$ is independent of n.

  - still outperforms MLE estimators.

  - worse than bal05 at small n, improves with large n but doesnt approach bal0.

- **regularisation is cheaper when the model is mispeccified.**

  - under correct specification, doubling λ roughly doubled the bias cost *(the gap between bal05 and bal0 was predictable from the KKT bound)*. under misspecification, the same λ increase costs less bias.

## Misspeccification + Overlap

```{r}
library(tidyverse)
library(kableExtra)

e2  <- read_csv("E2_results.csv.gz")
oxm <- read_csv("OVERLAP+MISSPECCIFICATION_results.csv.gz")
raw <- bind_rows(e2, oxm)

ests <- c("balnet0", "balnet05", "balnetcv", "glmcv_normal")
bal  <- c("balnet0", "balnet05", "balnetcv")

label_map <- c(
  balnet0 = "bal(\u03bb \u2192 0)", balnet05 = "bal(\u03bb = 0.05)",
  balnetcv = "bal(cv)", glmcv_normal = "glm(Hajek)"
)
color_map <- c(
  balnet0 = "#1a7a3a", balnet05 = "#2aa198",
  balnetcv = "#268bd2", glmcv_normal = "#dc322f"
)

summ <- raw |>
  filter(estimator %in% ests) |>
  group_by(n, p, misspec, outcome, overlap, estimator) |>
  summarise(
    m          = n(),
    bias       = mean(tau_hat),
    mcse_bias  = sd(tau_hat) / sqrt(m),
    empse      = sd(tau_hat),
    mcse_empse = empse / sqrt(2 * (m - 1)),
    rmse       = sqrt(mean(tau_hat^2)),
    mcse_rmse  = sd(tau_hat^2) / (2 * rmse * sqrt(m)),
    n_extreme  = sum(abs(tau_hat) > 10),
    .groups    = "drop"
  )

theme_sim <- theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"),
        panel.grid.minor = element_blank())
fmt_n <- labeller(n = \(x) paste("n =", format(as.numeric(x), big.mark = ",")))

plot_overlap <- function(outcome_val, title, which_ests = ests) {
  summ |>
    filter(estimator %in% which_ests, outcome == outcome_val,
           misspec == TRUE, overlap %in% c(0.25, 0.5, 0.75, 1)) |>
    mutate(estimator = factor(estimator, names(label_map), label_map)) |>
    ggplot(aes(overlap, rmse, colour = estimator)) +
    geom_line(linewidth = 0.8) + geom_point(size = 2.5) +
    facet_wrap(~ n, nrow = 1, labeller = fmt_n) +
    scale_colour_manual(values = setNames(color_map, label_map)) +
    scale_x_reverse(breaks = c(1, 0.75, 0.5, 0.25)) +
    expand_limits(y = 0) +
    labs(x = "Overlap scaler (c)", y = "RMSE", colour = "Estimator", title = title) +
    theme_sim
}

outcomes <- c(linear = "Linear", quad1 = "Quadratic", exp = "Exponential")

iwalk(outcomes, \(title, out) print(plot_overlap(out, title)))
iwalk(outcomes, \(title, out) print(plot_overlap(out, paste(title, "(Balancing Only)"), bal)))

# Delta RMSE tables (misspec - correct), n = 5000
for (out in c("linear", "exp", "quad1")) {
  summ |>
    filter(outcome == out, n == 5000) |>
    select(overlap, estimator, misspec, rmse) |>
    pivot_wider(names_from = misspec, values_from = rmse) |>
    mutate(delta = `TRUE` - `FALSE`) |>
    select(overlap, estimator, delta) |>
    pivot_wider(names_from = estimator, values_from = delta) |>
    arrange(desc(overlap)) |>
    kable(digits = 3,
          col.names = c("c", "bal0", "bal05", "balCV", "glmnet-Hajek"),
          caption = sprintf("Delta RMSE (misspec - correct) for %s (n = 5000, p = 50)", out)) |>
    kable_styling(full_width = FALSE) |>
    print()
}

# Delta RMSE plot
summ |>
  filter(n == 5000) |>
  select(outcome, overlap, estimator, misspec, rmse) |>
  pivot_wider(names_from = misspec, values_from = rmse) |>
  mutate(delta = `TRUE` - `FALSE`,
         outcome = factor(outcome, c("linear", "exp", "quad1")),
         estimator = factor(estimator, names(label_map), label_map)) |>
  ggplot(aes(overlap, delta, colour = estimator)) +
  geom_line(linewidth = 0.8) + geom_point(size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  facet_wrap(~ outcome) +
  scale_x_reverse(breaks = c(2, 1, 0.75, 0.5, 0.25)) +
  scale_colour_manual(values = setNames(color_map, label_map)) +
  labs(x = "Overlap scaler (c)", y = "\u0394RMSE (misspec \u2212 correct)", colour = "Estimator") +
  theme_sim

# Extreme estimates
summ |>
  filter(misspec == TRUE) |>
  mutate(type = if_else(grepl("balnet", estimator), "Balancing", "MLE")) |>
  group_by(type) |>
  summarise(extreme = sum(n_extreme), total = sum(m),
            pct = round(100 * extreme / total, 2))

# RMSE decomposed into bias and SD
summ |>
  filter(estimator %in% bal, n == 5000) |>
  pivot_longer(c(bias, empse), names_to = "metric") |>
  mutate(metric = factor(metric, c("bias", "empse"), c("|Bias|", "SD")),
         misspec = ifelse(misspec, "Misspecified", "Correct")) |>
  ggplot(aes(overlap, abs(value), colour = misspec)) +
  geom_line() + geom_point(size = 1.5) +
  facet_grid(metric ~ estimator + outcome, scales = "free_y") +
  labs(x = "Overlap (c)", y = NULL, colour = NULL) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom", strip.text.x = element_text(size = 7))

# Cell wins
summ |>
  filter(misspec == TRUE) |>
  group_by(outcome, n, overlap) |>
  slice_min(rmse, n = 1) |>
  ungroup() |>
  count(estimator, name = "cells_won")


# MSCE (simulation noise check -  possibly comeback later if relevent)

# Noisy RMSE cells (rel MCSE > 5%)
summ |>
  mutate(rel_mcse = mcse_rmse / rmse) |>
  filter(rel_mcse > 0.05) |>
  select(estimator, outcome, n, overlap, misspec, rel_mcse) |>
  arrange(desc(rel_mcse))

# Underpowered interaction effects (|z| < 3, |D_int| >= 0.05)
int_check |>
  filter(!estimator %in% c("glmnetcv", "oracle"),
         abs(z) < 3, abs(D_int) >= 0.05) |>
  select(estimator, outcome, n, overlap, D_int, z, m_needed) |>
  arrange(desc(abs(D_int)))
```

As expected, misspecification and overlap compound, and MLE methods are considerably outperformed by balancing methods. MLE RMSE increases substantially, while the balancing estimators remain mostly flat.

Under misspecification with imperfect overlap, balnet's attenuation bias is the cost of stability: RMSE stays bounded and improves with n, while MLE diverges significantly.

between methods, balancing protects against MLE's divergence. Within balnet, λ is a bias–variance dial where less is better except in situations with small n and weak overlap, where cv.balnet shines. In large samples, its bias floor is too high.

- **Against MLE, covariate balancing loss functions perform better across almost all cells, with advantage driven by λ selection.**

  - bal0 dominates every cell (45/45). Win count declines monotonically with λ: bal05 43/45, balcv 40/45. losses are a tuning problem, not a loss-function problem.
  - MLE only beats any balancing method at c=2, large n, with linear/exp outcomes (the 2–5 cells where glmnet's consistency overcomes balnet's attenuation bias). GLM never wins with a quadratic outcome.
  - Over all cells, concentrated around weak overlap, MLE produces 51 extreme IPW estimaties due to inversion; balancing methods produce 0. where an extreme estimate $|\hat{\tau}| > 10$ , where $\tau$ = 0. *(no need for ad hoc weight trimming or truncation for balnet)*

- **Effect of mispeccification compounds at lower overlap levels for both methods, but the interaction is substantially worse for MLE. (delta = RMSEmispec - RMSEcorrect)**

  - MLE delta increases substantially as overlap weakens. (max at delta = 2.45 for quadratic outcome).

    - *interaction effect grows with n for MLE, weak overlap accelerates K&S divergence . (eg: quad1 c=0.5: delta = 0.50 → 1.38 → 2.45 across n=500/1000/5000).*

    - *with strong overlap (c=2) divergence is mostly suppressed when mispeccified for linear and exponential outcomes.*

    - *compounding is clear for quadratic and linear outcomes; for exponential direction and ranking are clear, but the interaction is within MC noise (MCSE ≈ 0.13). Morris, White & Crowther (2019,* *Statistics in Medicine)*

  - balancing estimators stay bounded, and flat near 0. (max at delta = 0.34 with linear outcome). the small interaction that exists is balnets misspec residual bias growing as overlap weakens.

    - *Decomposing RMSE² into bias² + variance: misspecification shifts SD by at most 0.025 for bal0/bal05 — the penalty is almost entirely bias. balnetcv is the exception: SD falls under misspec as CV selects heavier λ, trading variance for bias.*

- **Within balnet, less regularization wins: bal0 is best in 36/45 cells; CV's λ is optimal where variance dominates**.

  - **balcv wins 6/45 cells:** weak overlap and low n for non linear outcomes (c≤0.5, n≤1000) These settings make CV's heavy shrinkage the right bias-variance trade. Otherwise, CV imposes an unnecessary bias floor.

  - **Path truncation for bal0 and bal05.** balnet0 = balnet05 exactly in 99.5% of reps at c=0.25/n=500 and 76.5% at n=1000. the path truncates above λ=0.05, as overlap is too weak to support that level of balance.

***notes:***

- ***reassess including bal10, does bal10 outperform balcv?***

*MSCE/simulation noise*

- *balnet estimates are clean everywhere (max relative MCSE 2.6%). no interaction cell falls below z=3. all rankings, signs, and gradients are stable.*

- *glm_hj point estimates are imprecise in misspec tails (rel MCSE up to 36%), but direction and ranking against balnet are never threatened — the gap between methods dwarfs the noise.*
