---
title: "Overlap (DGP2)"
format:
  html:
    mainfont: "Times New Roman"
editor_options:
  chunk_output_type: inline
---

# Setup

```{r}
#| message: false
#| warning: false
# Setup ------------------------------------------------------------
source("Analysis.R")
# DGP2, correct spec, p = 50, s × decay × covcor grid, overlap = 1, 1000 reps
df_sp <- load_sim("../results/sparsityDGP2.csv.gz")
summ_sp <- summarise_sim(df_sp) |>
  mutate(sf = factor(s, sort(unique(s))))
```

# INCREASE ACTIVE COVARIATES + DECAY

```{r}
dfm <- filter(summ_sp, covcor == "iid", decay == 1)   # flip covcor / decay for other tabs
sp_extra <- list(facet_wrap(~n, nrow = 1, labeller = lab_n), labs(x = "s"))

# RMSE | |Bias| vs s, faceted by n
for (oc in c("linear", "quad1")) print(panel(dfm, oc, x = sf, extra = sp_extra))

# bal0 vs oracle
for (oc in c("linear", "quad1")) print(
  panel(dfm, oc, x = sf, ests = c("balnet0", "oracle_hajek", "oracle_ht"),
        extra = sp_extra))

#- table -#
perf_table(dfm, cols = c("s", "n"),
           span_lab = \(x) paste0("s = ", x), span_desc = FALSE,
           group = "outcome", title = "Sparsity, iid, decay = 1 (c = 1)")

#- MSE decomp -#
for (oc in c("linear", "quad1")) print(
  decomp_plot(filter(dfm, outcome == oc), x = s, cols = n, xlab = "s",
              x_scale = scale_x_log10(breaks = c(4, 8, 16, 32, 50)),
              title = paste("MSE Decomposition (%) \u2014", oc_labs[oc])))

#- extreme estimates -#
extreme_heat(dfm, "glmnetcv_hajek", x = s, xlab = "s", x_desc = FALSE)
```

```{r}
dfm <- filter(summ_sp, covcor == "iid", decay == 0.5)   # flip covcor / decay for other tabs
sp_extra <- list(facet_wrap(~n, nrow = 1, labeller = lab_n), labs(x = "s"))

# RMSE | |Bias| vs s, faceted by n
for (oc in c("linear", "quad1")) print(panel(dfm, oc, x = sf, extra = sp_extra))

# bal0 vs oracle
for (oc in c("linear", "quad1")) print(
  panel(dfm, oc, x = sf, ests = c("balnet0", "oracle_hajek", "oracle_ht"),
        extra = sp_extra))

#- table -#
perf_table(dfm, cols = c("s", "n"),
           span_lab = \(x) paste0("s = ", x), span_desc = FALSE,
           group = "outcome", title = "Sparsity, iid, decay = 0.5 (c = 1)")

#- MSE decomp -#
for (oc in c("linear", "quad1")) print(
  decomp_plot(filter(dfm, outcome == oc), x = s, cols = n, xlab = "s",
              x_scale = scale_x_log10(breaks = c(4, 8, 16, 32, 50)),
              title = paste("MSE Decomposition (%) \u2014", oc_labs[oc])))

#- extreme estimates -#
extreme_heat(dfm, "glmnetcv_hajek", x = s, xlab = "s", x_desc = FALSE)
```

- BAL0 \> MLE \> BALCV:

  - baL0 unaffected increasing active covariates, reducing the rate of decay also doesnt seem to effect bal0.

  - bal0 unaffected by reducing sparsity.

  - bal0 by n = 10k, and active covariates = p = 50, reducing decay improves performance for MLE and balCV, they perform almost on par with bal0. bal05 and bal1q0 ared significantly degraded.

# CORRELATED COVARIATES+ DECAY

```{r}
dfm <- filter(summ_sp, covcor == "ar1", decay == 1)   # flip covcor / decay for other tabs
sp_extra <- list(facet_wrap(~n, nrow = 1, labeller = lab_n), labs(x = "s"))
# RMSE | |Bias| vs s, faceted by n
for (oc in c("linear", "quad1")) print(panel(dfm, oc, x = sf, extra = sp_extra))
# bal0 vs oracle
for (oc in c("linear", "quad1")) print(
  panel(dfm, oc, x = sf, ests = c("balnet0", "oracle_hajek", "oracle_ht"),
        extra = sp_extra))
#- table -#
perf_table(dfm, cols = c("s", "n"),
           span_lab = \(x) paste0("s = ", x), span_desc = FALSE,
           group = "outcome", title = "Sparsity, AR(1), decay = 1 (c = 1)")
#- MSE decomp -#
for (oc in c("linear", "quad1")) print(
  decomp_plot(filter(dfm, outcome == oc), x = s, cols = n, xlab = "s",
              x_scale = scale_x_log10(breaks = c(4, 8, 16, 32, 50)),
              title = paste("MSE Decomposition (%) \u2014", oc_labs[oc])))
#- extreme estimates -#
extreme_heat(dfm, "glmnetcv_hajek", x = s, xlab = "s", x_desc = FALSE)
```

```{r}
dfm <- filter(summ_sp, covcor == "ar1", decay == 0.5)   # flip covcor / decay for other tabs
sp_extra <- list(facet_wrap(~n, nrow = 1, labeller = lab_n), labs(x = "s"))
# RMSE | |Bias| vs s, faceted by n
for (oc in c("linear", "quad1")) print(panel(dfm, oc, x = sf, extra = sp_extra))
# bal0 vs oracle
for (oc in c("linear", "quad1")) print(
  panel(dfm, oc, x = sf, ests = c("balnet0", "oracle_hajek", "oracle_ht"),
        extra = sp_extra))
#- table -#
perf_table(dfm, cols = c("s", "n"),
           span_lab = \(x) paste0("s = ", x), span_desc = FALSE,
           group = "outcome", title = "Sparsity, AR(1), decay = 0.5 (c = 1)")
#- MSE decomp -#
for (oc in c("linear", "quad1")) print(
  decomp_plot(filter(dfm, outcome == oc), x = s, cols = n, xlab = "s",
              x_scale = scale_x_log10(breaks = c(4, 8, 16, 32, 50)),
              title = paste("MSE Decomposition (%) \u2014", oc_labs[oc])))
#- extreme estimates -#
extreme_heat(dfm, "glmnetcv_hajek", x = s, xlab = "s", x_desc = FALSE)
```

\`
