---
title: "Wang & Zubizarreta (2020) on the Wong & Chan design"
format: pdf
pdf-engine: xelatex
mainfont: Times New Roman
execute:
  echo: false
  warning: false
---

```{r setup}

library(dplyr)
dir <- paste0(here::here("CV Extension"), "/")
source(paste0(dir, "wc_grids.R"))   # rmse(), mcse(), cols, ltys, read_cell(), panel(),
# read_grid(), grid_fig(), pick_fig(), lambda_tab(), rmse_tab()
sbw <- data.table::fread(file = paste0(dir, "results/wc_sbw_v2.csv.gz"))
x   <- readRDS(paste0(dir, "results/wc_att_v2.rds"))
sig <- c(1, 10, 30, 100); cc <- 1:4          # path grids, drawn in the last section
ns  <- c(1000, 2500, 5000, 10000, 20000)
ks  <- c(10, 20, 65, 125); cb <- c(1, 3)
noise <- read_grid("wc_noise_s", sig, cc)
on    <- read_grid("wc_on_n", ns, cc)
basis <- read_grid("wc_basis_K", ks, cb)
```

**Setup:** their Wong & Chan simulation (n = 5000, 10 covariates, first and second moments balanced), two outcomes on the same draws, 1000 paired reps.

**Model A:** heavily confounded, heterogeneous effect.

**Model B:** barely confounded, no effect; behaves like low SNR.

Truth: 0 for the ATE; the sample ATT for the ATT (model A's ATT is about −17.7 and varies by draw).

# does approximate balance beat exact?

**Claim 1: under bad overlap:** exact balance often has no solution, and where it does the weights rest on a few observations, so approximate balance trades bias for variance and cuts ATT RMSE by half (pp. 94, 102, 103).

**Claim 2: under good overlap:** approximate and exact balance perform similarly (p. 102).

```{r sbw}
paper <- tibble::tribble(
  ~estimand, ~outcome, ~balance, ~rmse_paper,
  "ate", "A", "exact", 1.35, "ate", "A", "approx", 0.51,
  "ate", "B", "exact", 0.31, "ate", "B", "approx", 0.21,
  "att", "A", "exact", 0.72, "att", "A", "approx", 0.79,
  "att", "B", "exact", 0.26, "att", "B", "approx", 0.25)
sbw |>
  mutate(err = tau_hat - case_when(outcome == "B" ~ 0, estimand == "ate" ~ 0,
                                   TRUE ~ satt)) |>
  group_by(estimand, outcome, balance) |>
  summarise(rmse_ours = sqrt(mean(err^2)),
            bias_ours = mean(err),
            mcse_ours = sd(err^2) / sqrt(n()) / (2 * rmse_ours),
            .groups = "drop") |>
  left_join(paper, by = c("estimand", "outcome", "balance")) |>
  mutate(diff = rmse_ours - rmse_paper) |>
  arrange(estimand, outcome, desc(balance)) |>
  mutate(estimand = ifelse(duplicated(estimand), "", toupper(estimand))) |>
  select(estimand, outcome, balance, rmse_ours, rmse_paper, diff, mcse_ours,
         bias_ours) |>
  knitr::kable(digits = 2, booktabs = TRUE, linesep = "", align = "lllrrrrr",
               col.names = c("Estimand", "Outcome", "Balance", "RMSE ours",
                             "RMSE paper", "RMSE ours minus paper", "MCSE",
                             "Bias ours"))
```

Replication: model B and both exact arms match Table 4 within 0.35; model Aapproximate does not (2.4 to 2.7 times theirs). Our loop picks a median\$\\delta = 0.02\$; theirs must pick tighter, or the code differs. Supplementcode not yet checked.

**laim 2: under good overlap:** approximate and exact balance perform similarly (p. 102).

- **Model A:** fails. RMSE more than doubles (0.92 to 2.11), all bias. Paper: 10% worse.

- **Model B:** holds. RMSE falls 20% (0.29 to 0.23) - low SNR, perfect balancing fails.

The rule picks the same λ for both models; only the outcome, which it never sees, decides whether that λ helps.

# Selectors weights along the path

```{r att, fig.width=11, fig.height=5}
ok  <- vapply(x$res, \(r) is.null(r$err), logical(1))
res <- x$res[ok]
sbw <- subset(sbw, rep %in% which(ok) & estimand == "att")
lam_end <- vapply(res, `[[`, numeric(1), "lam_end")
reached <- colMeans(outer(lam_end, x$lam, "<="))
satt    <- vapply(res, `[[`, numeric(1), "satt")
lam_sel <- do.call(rbind, lapply(res, `[[`, "lam_sel"))
bind  <- \(o, k) do.call(rbind, lapply(res, \(r) r[[k]][o, ]))
truth <- \(o) if (o == "A") satt else 0
rmse  <- \(e) sqrt(colMeans(e^2))
mcse  <- \(e) apply(e^2, 2, sd) / sqrt(nrow(e)) / (2 * rmse(e))
sbw_err <- \(o, b) {
  z <- subset(sbw, outcome == o & balance == b)
  cbind(z$tau_hat - if (o == "A") z$satt else 0)
}
cols <- c(cv.bloss = "blue", cv.smd = "blue", cv.inf = "blue",
          boot.smd = "red", boot.inf = "red", alg1 = "black")
ltys <- c(4, 1, 2, 1, 2, 1)
layout(matrix(c(1, 2, 3, 3), 2, byrow = TRUE), heights = c(5, 1))
par(mar = c(4, 4, 3, 1))
tab <- do.call(rbind, lapply(c("A", "B"), \(o) {
  e_path <- bind(o, "est_path") - truth(o)
  r <- rmse(e_path); r[reached < 0.95] <- NA
  e_sel <- bind(o, "est_sel") - truth(o)
  e_end <- e_path[, max(which(!is.na(r))), drop = FALSE]
  e_opt <- e_path[, which.min(r), drop = FALSE]
  e_sbw <- list(exact = sbw_err(o, "exact"), approx = sbw_err(o, "approx"))
  plot(x$lam, r, log = "x", xlim = rev(range(x$lam)), type = "l", lwd = 2,
       ylim = range(c(r, rmse(e_sel), sapply(e_sbw, rmse)), na.rm = TRUE),
       xlab = "lambda (control-arm max SMD)", ylab = "RMSE vs sample ATT",
       main = paste("ATT, outcome model", o))
  abline(v = x$lam[which.min(r)], lty = 3, col = "gray40")
  abline(v = median(lam_end), lty = 3, col = "red")
  abline(h = rmse(e_sel), col = cols, lty = ltys)
  abline(h = sapply(e_sbw, rmse), col = "gray60", lty = c(2, 1), lwd = 2)
  row <- \(rule, e, lambda) data.frame(
    outcome = o, rule = rule, lambda = lambda, rmse = rmse(e), mcse = mcse(e))
  rbind(row("Best fixed lambda (oracle)", e_opt, x$lam[which.min(r)]),
        row("Exact balance (path floor)", e_end, median(lam_end)),
        row(c("cv.balnet, balance loss", "cv.balnet, mean SMD",
              "cv.balnet, max SMD", "cv.boot.balnet, mean SMD",
              "cv.boot.balnet, max SMD", "Their selector (Algorithm 1)"),
            e_sel, apply(lam_sel, 2, median)),
        row(c("sbw, exact", "sbw, their selector"),
            do.call(cbind, e_sbw), NA))
}))
par(mar = c(0, 0, 0, 0)); plot.new()
legend("center", c(names(cols), "sbw exact", "sbw approx"),
       col = c(cols, "gray60", "gray60"), lty = c(ltys, 2, 1), lwd = 2,
       horiz = TRUE, bty = "n")

show_model <- \(o, paper) {
  floor <- tab$rmse[tab$outcome == o & tab$rule == "Exact balance (path floor)"]
  tab |>
    filter(outcome == o) |>
    arrange(rmse) |>
    transmute(Method = rule,
              `median lambda` = ifelse(is.na(lambda), "",
                                       formatC(lambda, digits = 3, format = "f")),
              RMSE = sprintf("%.2f (%.3f)", rmse, mcse),
              `vs floor` = sprintf("%.2f", rmse / floor)) |>
    knitr::kable(align = "lrrr",
                 caption = sprintf("Model %s, ATT, 1000 reps. RMSE vs sample ATT (MCSE); vs floor = RMSE / exact-balance RMSE. Their Neg. Ent. row: %s.", o, paper))
}

show_model("A", "0.25 exact, 0.25 approx")


show_model("B", "0.25 exact, 0.25 approx")
```

Claims:

- \(1\) approximate balance beats approximate balance beats exact on RMSE

- \(2\) alg1, their bootstrapped lambda selector (boot.inf/boot.smd is balnets versions) lands at or near optimal lambda.

**Model A: (equivalent to high SNR)**

Both claims fail. Best λ is the floor (0.43).

Every rule relaxes and loses; theirs by 3.6× (1.54).

**Model B: (equivalent to low SNR)**

- Claim (1) holds: Best λ is 0.28 (0.19 oracle vs 0.32 at the floor).

- Claim (2) fails: every rule stops far short of the oracle (λ = 0.28, RMSE 0.19). Ranked:

1.  sbw, their selector (λ = 0.04 on sbw's scale, RMSE = 0.23)
2.  cv.balnet, mean SMD (λ = 0.032, RMSE = 0.24)
3.  cv.balnet, max SMD (λ = 0.036, RMSE = 0.24)
4.  Their selector, Algorithm 1 (λ = 0.010, RMSE = 0.26)
5.  cv.boot.balnet, mean SMD (λ = 0.004, RMSE = 0.27)
6.  cv.boot.balnet, max SMD (λ = 0.003, RMSE = 0.28)
7.  cv.balnet, balance loss (λ = 0.004, RMSE = 0.28)

The ranking is just the λ ranking, loser wins because B rewards not adjusting, on A it reverses. No rule can be right on both.
