---
title: "Overlap (DGP2)"
format:
  html:
    mainfont: "Times New Roman"
editor_options:
  chunk_output_type: inline
---

# Setup

```{r}
#| message: false
#| warning: false
# Setup ------------------------------------------------------------
source(here::here("R", "packages.R"))
source(here::here("R", "summarise.R"))
source(here::here("R", "plots.R"))
df_ov <- bind_rows(
  mutate(load_sim(here::here("results", "ipw", "dimensionDGP2.csv.gz")),        misspec = FALSE),
  mutate(load_sim(here::here("results", "ipw", "dimensionmisspecDGP2.csv.gz")), misspec = TRUE))
summ_dim <- summarise_sim(df_ov) |>
  mutate(pn = p / n, pnf = factor(pn, sort(unique(pn))))
```

# Dimensionality: correct spec

```{r}
dfm <- filter(summ_dim, !misspec)     # flip to misspec for the misspec tab
dim_extra <- list(facet_wrap(~n, nrow = 1, labeller = lab_n), labs(x = "p/n"))

# RMSE | |Bias| vs p/n, faceted by n
for (oc in outcome_cols) print(panel(dfm, oc, x = pnf, extra = dim_extra))

# bal0 vs oracle
for (oc in outcome_cols) print(
  panel(dfm, oc, x = pnf, ests = c("balnet0", "oracle_hajek", "oracle_ht"),
        extra = dim_extra))

#- table -#
perf_table(dfm, cols = c("pn", "n"),
           span_lab = \(x) paste0("p/n = ", x), span_desc = FALSE,
           group = "outcome", title = "Dimensionality (c = 1)")

#- MSE decomp -#
for (oc in outcome_cols) print(
  decomp_plot(filter(dfm, outcome == oc), x = pn, cols = n, xlab = "p/n",
              x_scale = scale_x_log10(breaks = c(0.1, 0.5, 1, 2)),
              title = paste("MSE Decomposition (%) \u2014", oc_labs[oc])))

#- extreme estimates -#
extreme_heat(dfm, "glmnetcv_hajek", x = pn, xlab = "p/n", x_desc = FALSE)
```

As expected, balancing estimators are more robust to high-dimensionality thn MLE estimation and Bal0 is still the best performing estimator, converging with bal05's path endpoint as $p/n$ increases. The only exception is for balnet.cv, which eventually underperforms MLE as dimensionality increases.

Under high dimensionality, Bal05 performs similarly to Bal0, especially in low sample sizes. Increases in sample size increase the edge bal0 has over bal05 and other balancing estimators.

The stability of bal10, and that bal0 and bal05 converge toward it, suggest that the feasible $\lambda_{endpoint}$ converges to \$ .05 \< \\lambda \\leq 0.1\$ as $p/n$ increases. Balnet.cv's performance suggest it consistently selects $\lambda > \lambda_{endpoint}$, and performs similarly or worse than MLE for $p/n ≥ 0.5$.

These results support that the balnet.cv cross validation criterion *(out-of-sample covariate balancing loss)* is suboptimal.

**new quick notes:**

- bal0 still best performer, relative performance increases with n.

- bal0\>bal05 ahve the highest share of variance in their MSE decomposition.

**old notes:**

**1: Bal0 is still the strongest estimator\***,

- loses its edge in lower samples, where $n=500$ bal0 converges with bal05 by $p/n =.5$ , and $p/n = 1$ when $n=1000$.

  - these are the two estimators the with the least bias dominated RMSE.

    - *bal05 consistently beats glm.*

    - **for all estimators,** RMSE becomes more bias driven as $p/n$ increases, except for bal10 who's bias share is flat. glmnet/bal10 are the most bias dominated estimators, and start off almost completely bias dominated in linear specifications.

  - as $p/n$ increases bal0 loses its consistency against the normalized oracle as regularisation binds bal0's allowed balance level, $\lambda$, higher.

    - consistency result survives for quadratic specifications n=500, but for n=1000 only survive upto $p/n \approx 1$. Against non-normalized oracle, consistency results survive both nonlinear outcomes.

    - *now loses to the normalized oracle under linear specifications when* $p/n \ge 1$*. it also loses to the normalized oracle under exponential specifications at* $n=1000$ $p/n \ge 1$.

**2: bal10 has stable variance and bias across all** $p/n$ **levels.**

- bal10's bias is stable across $p/n$ as KKT conditions bound the maximum covariate imbalance to $\lambda=.1$ (Sverdrup and Hastie 2026, eq. 10).

- bal10's flatness and that bal0 and bal05 converge towards it as $p/n$ increases, indicating the feasible $\lambda_{endpoint}$ converges with $p/n$ increases to somewhere close (but below) $\lambda = .1$ *(can log regularisation path in future to determine where).*

  - for most $p/n ≥ 1$ the $\lambda_{endpoint}$ exceeds 0.05, so bal05 and bal0 produce near identical results *(small bal0 wins at some* $p/n = 1$*, could be sampling noise).\**

**3: at** $p/n \ge 0.5$ **Balnet.cv and MLE and are consistently the worst performers.** their comparitve performance depends on outcome complexity.

- balcv degrades *faster* than glmnet as $p/n$ grows and crosses below glm_hj . The loss function survives dimensionality, but the CV rule doesn't. **strong evidence against balnet.cv.**

  - at $p/n$ = 0.1, balcv generally outperforms bal10 and glm.

    - *CV still adapts at p/n = 0.1 but detaches by p/n = 0.5. **FUTURE:** logging λ_cv across the path will map the transition.*

- Under high dimensionality, once $p/n = 1$ MLE outperforms balnet.cv decicivly for linear and moderately for quadratic outcomes, approximatley tied under exponential outcomes.

- increases in $n$ reduce their variance, causing them to converge towards bal10.

**4: larger samples reduces\* RMSE for all estimators across all** $p/n$

- \*for bal10, variance reduces with $n$ while bias slightly increases. RMSE is also flat across the linear specification where RMSE is constantly $\approx$ 0.25 .

**5: outcome complexity effects relative bias/RMSE differently.**

- **exponential:** reduces bias/RMSE for all estimators *(the one exception is bal0 at* $p/n =$ *.1)* , ranking remains the same.

- **quadratic:** increases RMSE at low $p/n$, but **reduces it for the most biased estimators (balcv, glm) at high** $\mathbf{p}/\mathbf{n}$, where the bias reduction outweighs the higher outcome variance.

  - *at* $n=1000$ *bal.cv equivelent to glm.cv but loses to bal10 at* $p/n = .5$

```{r}
# Extended grid: merge p/n ∈ {4, 8} (linear + quad1 only) with correct-spec results
df_pm <- bind_rows(
  filter(df_ov, !misspec, outcome != "exp"),
  mutate(load_sim(here::here("results", "ipw", "PoverMDGP2.csv.gz")), misspec = FALSE))
summ_pm <- summarise_sim(df_pm) |>
  mutate(pn = p / n, pnf = factor(pn, sort(unique(pn))))

dfm <- summ_pm
dim_extra <- list(facet_wrap(~n, nrow = 1, labeller = lab_n), labs(x = "p/n"))

# RMSE | |Bias| vs p/n, faceted by n
for (oc in c("linear", "quad1")) print(panel(dfm, oc, x = pnf, extra = dim_extra))

# bal0 vs oracle
for (oc in c("linear", "quad1")) print(
  panel(dfm, oc, x = pnf, ests = c("balnet0", "oracle_hajek", "oracle_ht"),
        extra = dim_extra))

#- table -#
perf_table(dfm, cols = c("pn", "n"),
           span_lab = \(x) paste0("p/n = ", x), span_desc = FALSE,
           group = "outcome", title = "Dimensionality, extended p/n (c = 1)")

#- MSE decomp -#
for (oc in c("linear", "quad1")) print(
  decomp_plot(filter(dfm, outcome == oc), x = pn, cols = n, xlab = "p/n",
              x_scale = scale_x_log10(breaks = c(0.1, 0.5, 1, 2, 4, 8)),
              title = paste("MSE Decomposition (%) \u2014", oc_labs[oc])))

#- extreme estimates -#
extreme_heat(dfm, "glmnetcv_hajek", x = pn, xlab = "p/n", x_desc = FALSE)
```

# Misspecification + Dimensionality

```{r}
dfm <- filter(summ_dim, misspec)
dim_extra <- list(facet_wrap(~n, nrow = 1, labeller = lab_n), labs(x = "p/n"))

# RMSE | |Bias| vs p/n, faceted by n
for (oc in outcome_cols) print(
  panel(dfm, oc, x = pnf, extra = dim_extra,
        title = paste(oc_labs[oc], "\u2014 misspecified")))

# bal0 vs oracle
for (oc in outcome_cols) print(
  panel(dfm, oc, x = pnf, ests = c("balnet0", "oracle_hajek", "oracle_ht"),
        extra = dim_extra, title = paste(oc_labs[oc], "\u2014 misspecified")))

#- table -#
perf_table(dfm, cols = c("pn", "n"),
           span_lab = \(x) paste0("p/n = ", x), span_desc = FALSE,
           group = "outcome", title = "Dimensionality, misspecified (c = 1)")

#- MSE decomp -#
for (oc in outcome_cols) print(
  decomp_plot(filter(dfm, outcome == oc), x = pn, cols = n, xlab = "p/n",
              x_scale = scale_x_log10(breaks = c(0.1, 0.5, 1, 2)),
              title = paste("MSE Decomposition (%), misspecified \u2014", oc_labs[oc])))

#- extreme estimates -#
extreme_heat(dfm, "glmnetcv_hajek", x = pn, xlab = "p/n", x_desc = FALSE,
             title = sprintf("GLM (Hajek), misspecified: extreme estimates (|\u03c4\u0302| > 10)"))
```

```{r}

```
