---
title: "Working Analysis"
format:
  html:
    mainfont: "Times New Roman"
editor_options: 
  chunk_output_type: inline
---

# Setup

```{r}
setwd("~/Thesis 2026/Masters_Thesis")
#| message: false
#| warning: false
library(tidyverse)   # ggplot2, dplyr, tidyr, purrr, stringr
library(ggh4x)       # nested facet strips (the |Bias|/SD × estimator figure)
library(patchwork)   # RMSE | |Bias| layout
library(gt)          # tables
library(kableExtra)  # tables

options(readr.show_col_types = FALSE)

# ── 1. Summaries: Monte Carlo performance measures (Morris, White & Crowther 2019) ──
# Keeps ALL estimators; subsetting happens in the figure calls, never here.
# `by` is a parameter so sections with different design columns (e.g. no `overlap`)
# can drop them: summarise_sim(df, by = c("n","p","outcome","estimator")).

ummarise_sim <- function(data,
                         by = c("n","p","misspec","covcor","outcome",
                                "overlap","signs","estimator")) {
  by <- intersect(by, names(data))   # design cols absent from `data` drop out silently
  data |>
    group_by(across(all_of(by))) |>
    summarise(
      m          = dplyr::n(),
      bias       = mean(tau_hat),
      empse      = sd(tau_hat),
      rmse       = sqrt(mean(tau_hat^2)),
      mcse_bias  = empse / sqrt(m),
      mcse_empse = empse / sqrt(2 * (m - 1)),
      mcse_rmse  = sd(tau_hat^2) / (2 * rmse * sqrt(m)),
      n_extreme  = sum(abs(tau_hat) > 10),
      .groups    = "drop"
    )
}

# ── 2. One label map + one palette (Okabe-Ito) + reusable estimator subsets ──
lab_est <- c(balnet0="BalNet (\u03bb=0)", balnet05="BalNet (\u03bb=.05)",
             balnet10="BalNet (\u03bb=.10)", balnetcv="BalNet (CV)",
             glmcv_normal="GLM (Hajek)", glmnetcv="GLMNet (CV)",
             normalised_oracle="Norm. Oracle", oracle ="oracle")
pal_est <- c(balnet0="#009E73", balnet05="#56B4E9", balnet10="#E69F00",
             balnetcv="#0072B2", glmcv_normal="#B22222", glmnetcv="#CC79A7",
             normalised_oracle="grey55", oracle = "#404040")
est_main <- c("balnet0","balnet05","balnet10","balnetcv","glmcv_normal")
est_bal  <- c("balnet0","balnet05","balnet10","balnetcv")  
oc_labs  <- c(linear="Linear", quad1="Quadratic", exp="Exponential")
sign_labs <- c(pos = "All +", neg = "All \u2212", mixed = "Alternating",
               ks = "K&S (+,\u2212,+,+)")

pal_spec <- c(Correct = "#009E73", Misspecified = "#B22222")   # specification contrast (spec_grid)



# ── 3. Shared theme + reusable scale/facet pieces ──
theme_sim <- theme_minimal(base_size = 12) +
  theme(legend.position  = "bottom",
        legend.title     = element_blank(),
        strip.background = element_rect(fill = "grey90", colour = NA),
        strip.text       = element_text(face = "bold"),
        panel.grid.minor = element_blank(),
        axis.text.x      = element_text(angle = 45, hjust = 1))

scale_x_n <- scale_x_log10(breaks = c(500,1000,5000,10000,50000),
                           labels = c("500","1k","5k","10k","50k"))
as_c   <- function(x) factor(paste0("c = ", x),
                             levels = paste0("c = ", sort(unique(x), decreasing = TRUE)))
lab_n  <- as_labeller(\(x) paste0("n = ", format(as.numeric(x), big.mark = ",")))
n_lab  <- c("100"="100","200"="200","250"="250","400"="400",
            "500"="500","750"="750","800"="800",
            "1000"="1k","2000"="2k","3000"="3k","4000"="4k","5000"="5k",
            "10000"="10k","50000"="50k")


# ── 4. Core line plot: metric vs x, coloured by estimator ──
# y/ylab + x + ests are arguments; x-scale and facets are appended by the caller.
line_plot <- function(data, y, ylab, x = n, ests = est_main, from_zero = TRUE) {
  ests <- intersect(names(lab_est), ests)                 # canonical order, no ghost keys
  p <- data |>
    filter(estimator %in% ests) |>
    ggplot(aes({{ x }}, {{ y }}, colour = estimator, group = estimator)) +
    geom_line(linewidth = 0.7) + geom_point(size = 1.8) +
    scale_colour_manual(values = pal_est, breaks = ests,
                        labels = \(b) unname(lab_est[b])) +
    labs(y = ylab, colour = NULL) + theme_sim
  if (from_zero) p + expand_limits(y = 0) else p
}

# ── 5. Twin panel: RMSE | |Bias|, one shared legend ──
panel <- function(data, oc, x = n, ests = est_main, extra = NULL,
                  title = oc_labs[oc], from_zero = TRUE, compact = FALSE) {
  d <- filter(data, outcome == oc)
  drop0 <- \(x) sub("^(-?)0\\.", "\\1.", x)
  yl <- if (compact) NULL else "RMSE"
  yr <- if (compact) NULL else "|Bias|"
  l <- line_plot(d, rmse,      yl, x = {{ x }}, ests = ests, from_zero = from_zero) + extra +
    theme(legend.position = "none")
  r <- line_plot(d, abs(bias), yr, x = {{ x }}, ests = ests, from_zero = from_zero) + extra +
    theme(legend.position = "none")
  if (compact) {
    l <- l + ggtitle("RMSE") + scale_y_continuous(labels = drop0)
    r <- r + ggtitle("|Bias|") + scale_y_continuous(labels = drop0)
  }
  (l | r) + plot_layout(guides = "collect") +
    plot_annotation(title = title) & theme(legend.position = "bottom")
}

# ── 6. MSE decomposition: Bias² / Variance share, stacked area ──
pal_comp <- c("Variance" = "#377eb8", "Bias\u00b2" = "#e41a1c")

decomp_plot <- function(data, ests = est_main, x = n, cols = outcome,
                        x_scale = scale_x_n, xlab = "n",
                        col_lab = labeller(outcome = oc_labs, n = lab_n),
                        title = NULL) {
  keep <- intersect(names(lab_est), ests)
  data |>
    filter(estimator %in% keep) |>
    mutate(denom    = bias^2 + empse^2,
           bias_sq  = bias^2  / denom,
           variance = empse^2 / denom,
           estimator = factor(lab_est[estimator], levels = lab_est[keep])) |>
    pivot_longer(c(variance, bias_sq), names_to = "component", values_to = "share") |>
    mutate(component = factor(component, c("variance","bias_sq"),
                              c("Variance","Bias\u00b2"))) |>
    ggplot(aes({{ x }}, share, fill = component)) +
    geom_area() +
    facet_grid(rows = vars(estimator), cols = vars({{ cols }}), switch = "y",
               labeller = col_lab) +
    x_scale +
    scale_y_continuous(breaks = c(1), labels = scales::percent) +
    scale_fill_manual(values = pal_comp) +
    labs(x = xlab, y = "Share of MSE", fill = NULL, title = title) +
    theme_sim +
    theme(strip.text.y.left = element_text(angle = 0))
}

# ── 7. Performance table: estimators × design columns, ordered by mean metric ──
# cols="n" → flat.  cols=c("overlap","n") → c-spanners.
# group="outcome" → stacks all outcomes in one table with row groups.
perf_table <- function(data, ests = est_main, cols = "n",
                       top = "rmse", bottom = "abs_bias",
                       digits = 2, order_by = "rmse",
                       lab = lab_est, n_lab_tbl = n_lab,
                       span_lab = \(x) paste0("c = ", x), span_desc = TRUE,
                       title = "", subtitle = NULL,
                       group = NULL, group_lab = NULL) {
  stopifnot(length(cols) %in% 1:2,
            c(top, bottom, order_by) %in% c("rmse","bias","abs_bias","empse"))
  keep <- intersect(names(lab), ests)
  cvar <- cols[length(cols)]
  svar <- if (length(cols) == 2L) cols[[1]] else NULL
  pm   <- c(rmse="RMSE", bias="Bias", abs_bias="|Bias|", empse="EmpSE")
  f    <- paste0("%.", digits, "f")
  grp  <- !is.null(group)
  
  d <- data |> filter(estimator %in% keep) |> mutate(abs_bias = abs(bias)) |> mutate(
    Estimator = factor(unname(lab[estimator]), levels = unname(lab[keep])),
    cell = sprintf(paste0(f, "<br><small>(", f, ")</small>"), .data[[top]], .data[[bottom]]),
    avg  = ave(.data[[order_by]], estimator, FUN = mean),
    .col = if (is.null(svar)) paste0(cvar, "=", .data[[cvar]])
    else paste0(svar, "=", .data[[svar]], "|", cvar, "=", .data[[cvar]]))
  if (grp) {
    gv <- as.character(data[[group]])                 # as.character: a factor would index by integer code
    gl <- group_lab %||% switch(group, outcome = oc_labs, signs = sign_labs,
                                setNames(nm = sort(unique(gv))))
    gl <- gl[names(gl) %in% gv]                       # drop unused levels
  d  <- mutate(d, Group = factor(gl[as.character(.data[[group]])], levels = unname(gl)))
}
  
  cv  <- sort(unique(d[[cvar]]))
  sv  <- if (is.null(svar)) NULL else sort(unique(d[[svar]]), decreasing = span_desc)
  key <- if (is.null(svar)) paste0(cvar, "=", cv)
  else c(t(outer(sv, cv, \(s, c) paste0(svar, "=", s, "|", cvar, "=", c))))
  
  id_vars <- c(if (grp) "Group", "Estimator", ".col", "cell", "avg")
  wide <- d |> distinct(across(all_of(id_vars))) |>
    pivot_wider(names_from = .col, values_from = cell)
  wide <- if (grp) arrange(wide, Group, avg) else arrange(wide, avg)
  present <- intersect(key, names(wide))
  wide <- wide |> select(any_of("Group"), Estimator, all_of(present))
  
  inner_lab <- \(x) { x <- sub("^[^=]*=", "", sub(".*\\|", "", x))
  if (is.null(n_lab_tbl)) x else unname(ifelse(x %in% names(n_lab_tbl), n_lab_tbl[x], x)) }
  
  g <- if (grp) gt(wide, groupname_col = "Group") else gt(wide)
  g <- g |> tab_header(title = title,
                       subtitle = md(subtitle %||% paste0(pm[[top]], " (", pm[[bottom]], ")")))
  for (s in sv) g <- tab_spanner(g, span_lab(s), starts_with(paste0(svar, "=", s, "|")))
  g |>
    cols_label_with(all_of(present), inner_lab) |>
    fmt_markdown(all_of(present)) |> cols_align("center", all_of(present)) |>
    tab_style(cell_text(weight = "bold"), cells_body(Estimator)) |>
    opt_table_lines("default") |>
    tab_source_note(md(sprintf("_Top %s / bottom %s; sorted by mean %s._",
                               pm[[top]], pm[[bottom]], pm[[order_by]])))
}


# ── 8. Convergence / path-truncation table ──
# % reps where ALL selected estimators coincide (range < tol).
# k=2: bal0 ≡ bal05.  k≥3: full-path collapse.  Raw replicates, not summaries.
converge_table <- function(data, ests, by = c("overlap","n"),
                           cols = by[length(by)], tol = 1e-9,
                           outcome_name = NULL, overlap_vals = NULL,
                           n_lab_tbl = n_lab, title = NULL) {
  stopifnot(length(ests) >= 2, length(cols) == 1, cols %in% by)
  d <- filter(data, estimator %in% ests,
              if (is.null(outcome_name)) TRUE else outcome == outcome_name,
              if (is.null(overlap_vals)) TRUE else overlap %in% overlap_vals)
  
  wide <- d |>
    group_by(across(all_of(c(by, "sim")))) |>
    summarise(hit = (max(tau_hat) - min(tau_hat)) < tol, .groups = "drop") |>
    group_by(across(all_of(by))) |>
    summarise(pct = sprintf("%.1f%%", mean(hit) * 100), .groups = "drop") |>
    pivot_wider(names_from = all_of(cols), values_from = pct, names_prefix = paste0(cols, "="))
  
  ord  <- grep(paste0("^", cols, "="), names(wide), value = TRUE)
  ord  <- ord[order(as.numeric(sub(".*=", "", ord)))]
  wide <- select(wide, all_of(setdiff(by, cols)), all_of(ord))
  if ("overlap" %in% names(wide)) names(wide)[names(wide) == "overlap"] <- "c"
  
  g <- gt(wide)
  if (!is.null(title)) g <- tab_header(g, title = md(title))
  g |>
    cols_label_with(all_of(ord), \(x) { v <- sub(".*=", "", x)
    if (is.null(n_lab_tbl)) v else unname(ifelse(v %in% names(n_lab_tbl), n_lab_tbl[v], v)) }) |>
    cols_align("center")
}

# ── 9. Misspecification helpers ──
# ΔRMSE and Δ|Bias| = misspecified − correct at matching (n, outcome, overlap, estimator).
delta_ms <- function(summ, ests = est_main) {
  summ |>
    filter(estimator %in% ests) |>
    mutate(absbias = abs(bias)) |>
    select(n, outcome, overlap, estimator, misspec, rmse, absbias) |>
    pivot_wider(names_from = misspec, values_from = c(rmse, absbias)) |>
    mutate(d_rmse = rmse_TRUE - rmse_FALSE,
           d_bias = absbias_TRUE - absbias_FALSE)
}

delta_rmse <- function(summ, ests = est_main)
  delta_ms(summ, ests) |> transmute(n, outcome, overlap, estimator, delta = d_rmse)

# §9 — compact ΔRMSE | Δ|Bias| twin, faceted by n, walked over outcomes
delta_panel <- function(data, oc, ests = est_main, ov = c(0.25, 0.5, 0.75, 1),
                        title = oc_labs[oc]) {
  drop0 <- \(x) sub("^(-?)0\\.", "\\1.", x)
  d <- filter(data, outcome == oc, overlap %in% ov)
  shared <- list(
    geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4),
    scale_x_reverse(breaks = ov),
    facet_wrap(~n, labeller = lab_n),
    scale_y_continuous(labels = drop0))
  l <- line_plot(d, d_rmse, NULL, x = overlap, ests = ests, from_zero = FALSE) +
    ggtitle("\u0394RMSE") + shared + theme(legend.position = "none")
  r <- line_plot(d, d_bias, NULL, x = overlap, ests = ests, from_zero = FALSE) +
    ggtitle("\u0394|Bias|") + shared + theme(legend.position = "none")
  (l | r) + plot_layout(guides = "collect") +
    plot_annotation(title = title) & theme(legend.position = "bottom")
}

# Style-(a): |Bias| and SD vs overlap, coloured by specification (Correct vs Misspecified).
spec_grid <- function(data, ests = est_bal, at_n = 5000, ov = c(0.25, 0.5, 0.75, 1, 2)) {
  ests <- intersect(names(lab_est), ests)
  data |>
    filter(estimator %in% ests, n == at_n, overlap %in% ov) |>
    pivot_longer(c(bias, empse), names_to = "metric") |>
    mutate(metric    = factor(metric, c("bias","empse"), c("|Bias|","SD")),
           spec      = ifelse(misspec, "Misspecified", "Correct"),
           estimator = factor(lab_est[estimator], levels = lab_est[ests]),
           outcome   = factor(outcome, names(oc_labs), oc_labs)) |>
    ggplot(aes(overlap, abs(value), colour = spec)) +
    geom_line() + geom_point(size = 1.5) +
    facet_grid(metric ~ estimator + outcome, scales = "free_y") +
    scale_x_continuous(breaks = c(0.25, 0.5, 1, 2), labels = c(".25", ".5", "1", "2")) +
    scale_colour_manual(values = pal_spec) +
    labs(x = "Overlap (c)", y = NULL, colour = NULL) +
    theme_bw(base_size = 10) +
    theme(legend.position = "bottom",
          strip.text.x = element_text(size = 7),
          axis.text.x  = element_text(angle = 45, hjust = 1, size = 7))
}


# ── 10. Extreme-estimate heatmap: reps with |tau_hat| > 10, overlap × n, faceted by outcome ──
# Tile colour + numerator = `est`'s extreme count per cell; denominator = `est`'s total across
# the whole figure. Labels only where the cell has ≥1 (empty cells left blank).
extreme_heat <- function(data, est = "glmcv_normal", ov = NULL,
                         high = "#B22222", title = NULL) {
  d <- data |>
    filter(estimator == est, if (is.null(ov)) TRUE else overlap %in% ov)
  total <- sum(d$n_extreme)                       # est's extreme estimates across the figure
  cells <- d |>
    group_by(outcome, n, overlap) |>
    summarise(extreme = sum(n_extreme), .groups = "drop") |>
    mutate(fill_val = if_else(extreme > 0, extreme, NA_real_),
           outcome  = factor(outcome, names(oc_labs), oc_labs),
           n        = factor(n, sort(unique(n))),
           overlap  = factor(overlap, sort(unique(overlap), decreasing = TRUE)))
  ggplot(cells, aes(overlap, n, fill = fill_val)) +
    geom_tile(colour = "white", linewidth = 0.6) +
    geom_text(aes(label = if_else(extreme > 0, sprintf("%.0f/%.0f", extreme, total), "")),
              size = 3) +
    facet_wrap(~outcome) +
    scale_fill_gradient(low = "white", high = high, name = "Extreme\n(count)", na.value = "white") +
    scale_y_discrete(labels = n_lab) +
    labs(x = "Overlap (c)", y = "n",
         title = title %||% sprintf("%s: extreme estimates (|\u03c4\u0302| > 10)",
                                    unname(lab_est[est]))) +
    theme_sim +
    theme(panel.grid = element_blank(), legend.position = "right")
}




```

```{r}

```

## AR1: Correct Specification

```{r}
# ── E1: Correctly specified DGP (strong overlap, c = 1) ──
df_e1 <- read_csv("ar1GDP1.csv.gz")

summ_e1 <- summarise_sim(df_e1) |> filter(!misspec, overlap == 1)
est_e1  <- c(est_main, "normalised_oracle")     # 5 main estimators + normalised oracle

# RMSE | |Bias| vs n, per outcome
walk(c("linear", "quad1", "exp"),
     \(oc) perf_table(filter(summ_e1, outcome == oc), ests = est_e1,
                      group = "signs", title = paste("E1 —", oc)))

# RMSE (|Bias|) tables, per outcome
perf_table(summ_e1, ests = est_e1, group = "outcome", title = "E1: Correctly Specified")

# MSE decomposition (Bias²/Variance share), estimators × outcome (no oracle)
print(decomp_plot(summ_e1, ests = est_main,
                  title = "MSE Decomposition (%) \u2014 Correct spec (c = 1)"))
```

- **Under linear outcomes:** MLE and Set $\lambda \leq .05$ balancing estimators perform better under AR(1) correlation than when iid is assumed, both relatively and in absolute magnitude. Only Balnet (CV) is degraded ie:

  - at n=500 Bal0 estimated RMSE is 0.11, and 0.13 for iid and ar1 respecitvely. for bias, 0.01 vs 0.02.

  - at n=1000 normalized glmnet has RMSE 0.17 vs 0.2 with Bias 0.15 vs .18, ar1 and iid respectively.

  - CVbalnet: n=500, RMSE = 0.28 vs 0.24, Bias 0.25 vs 0.18, ar1 and iid respectively.

  - gains are stronger for balancing estimators relative to MLE (bal05 outperforms MLE up to n=5000).

- Under quadratic outcomes: AR(1) correlation degrades, set Set $\lambda \leq .05$ balancing estimators perform better relative to MLE.

  - for bal0 and bal05 RMSE and bias absolute magnitude vs their iid equivilent are only moderately degraded, and could be sampling artifacts.

  - MLE is degraded vs when iid at $n \leq 1000$

- exponential results are effectively the same as under linear outcomes.

```{r}
# ── E1: Correctly specified DGP (strong overlap, c = 1) ──
df_e1 <- read_csv("DGP2correctSIGN.csv.gz")

summ_e1 <- summarise_sim(df_e1) |> filter(!misspec, overlap == 1)
est_e1  <- c(est_main, "normalised_oracle")     # 5 main estimators + normalised oracle


# RMSE | |Bias| vs n, per outcome
walk(c("linear","quad1","exp"),
     \(oc) print(panel(summ_e1, oc, ests = est_e1, extra = scale_x_n)))

# RMSE (|Bias|) tables, per outcome
perf_table(summ_e1, ests = est_e1, group = "outcome", title = "E1: Correctly Specified")

# MSE decomposition (Bias²/Variance share), estimators × outcome (no oracle)
print(decomp_plot(summ_e1, ests = est_main,
                  title = "MSE Decomposition (%) \u2014 Correct spec (c = 1)"))
```
