---
title: "Working Analysis"
format:
  html:
    mainfont: "Times New Roman"
editor_options: 
  chunk_output_type: inline
---

# Setup

```{r}

library(dplyr)
library(tidyr)      # pivot_wider/longer
library(ggplot2)
library(patchwork)  # (l | r) panels
library(gt)
library(scales)     # percent labels
library(tibble)     # tribble (lam_meta)

# pin dplyr verbs against masking (MASS, stats, etc.)
select <- dplyr::select
filter <- dplyr::filter

options(readr.show_col_types = FALSE)


# ============================================================================
# analysis_setup_patch.R
# PART A: replace the matching blocks in the existing setup (sections 1-2).
# PART B: append as new sections 11-16.
# Every existing figure/table call (E1-E5 scripts) runs unchanged on old AND
# new files; the new logging columns are only touched by sections 11-16.
# ============================================================================

# ── A0. New DGP logs diagnostics as rows of `estimator`; name them once ──
diag_stats <- c("lam_balcv1","lam_balcv0","lam_end1","lam_end0","lam_glmcv",
                "trunc05","nnz_balcv1","nnz_balcv0","nnz_glm",
                "smd1_05","smd0_05","smd1_cv","smd0_cv","err")

# ── A1. Section 1 replacement: summarise_sim ──
# Changes, all no-ops on old files:
#   (i)   diagnostic rows dropped — never summarised as estimators
#   (ii)  s, strength join the default `by` (intersect() drops them when the
#         columns are absent, so old summaries are identical)
#   (iii) n_extreme gains na.rm = TRUE (old files screened at 0 NAs)
#   (iv)  new n_na column for the NA screen; bias/empse/rmse still propagate
#         NA loudly by design — screen first, as usual
summarise_sim <- function(data,
                          by = c("n","p","s","misspec","covcor","outcome",
                                 "overlap","signs","strength","estimator")) {
  by <- intersect(by, names(data))
  data |>
    filter(!estimator %in% diag_stats) |>
    group_by(across(all_of(by))) |>
    summarise(
      m          = dplyr::n(),
      bias       = mean(tau_hat),
      empse      = sd(tau_hat),
      rmse       = sqrt(mean(tau_hat^2)),
      mcse_bias  = empse / sqrt(m),
      mcse_empse = empse / sqrt(2 * (m - 1)),
      mcse_rmse  = sd(tau_hat^2) / (2 * rmse * sqrt(m)),
      n_extreme  = sum(abs(tau_hat) > 10, na.rm = TRUE),
      n_na       = sum(is.na(tau_hat)),
      .groups    = "drop"
    )
}

# ── A2. Section 2 replacement: label map + palette gain balnetrate ──
# est_main / est_bal untouched, so every existing call renders identically.
lab_est <- c(balnet0="BalNet (\u03bb=0)", balnet05="BalNet (\u03bb=.05)",
             balnet10="BalNet (\u03bb=.10)", balnetcv="BalNet (CV)",
             balnetrate="BalNet (\u221a(log p/n))",
             glmcv_normal="GLM (Hajek)", glmnetcv="GLMNet (CV)",
             normalised_oracle="Norm. Oracle", oracle="oracle")
pal_est <- c(balnet0="#009E73", balnet05="#56B4E9", balnet10="#E69F00",
             balnetcv="#0072B2", balnetrate="#D55E00",
             glmcv_normal="#B22222", glmnetcv="#CC79A7",
             normalised_oracle="grey55", oracle="#404040")
est_main <- c("balnet0","balnet05","balnet10","balnetcv","glmcv_normal")
est_bal  <- c("balnet0","balnet05","balnet10","balnetcv")
est_rate <- c("balnet0","balnetcv","balnetrate","glmcv_normal")  # rate-\u03bb contrast

lab_signs <- c(ks = "K&S (+,\u2212,+,+)", pos = "All positive", mixed = "Alternating")

# Sections 3-10 unchanged.

# ── 3. Shared theme + reusable scale/facet pieces ──
theme_sim <- theme_minimal(base_size = 12) +
  theme(legend.position  = "bottom",
        legend.title     = element_blank(),
        strip.background = element_rect(fill = "grey90", colour = NA),
        strip.text       = element_text(face = "bold"),
        panel.grid.minor = element_blank(),
        axis.text.x      = element_text(angle = 45, hjust = 1))

scale_x_n <- scale_x_log10(breaks = c(500,1000,5000,10000,50000),
                           labels = c("500","1k","5k","10k","50k"))
as_c   <- function(x) factor(paste0("c = ", x),
                             levels = paste0("c = ", sort(unique(x), decreasing = TRUE)))
lab_n  <- as_labeller(\(x) paste0("n = ", format(as.numeric(x), big.mark = ",")))
n_lab  <- c("100"="100","200"="200","250"="250","400"="400",
            "500"="500","750"="750","800"="800",
            "1000"="1k","2000"="2k","3000"="3k","4000"="4k","5000"="5k",
            "10000"="10k","50000"="50k")
oc_labs <- c(linear = "Linear", quad1 = "Quadratic", exp = "Exponential")


# ── 4. Core line plot: metric vs x, coloured by estimator ──
# y/ylab + x + ests are arguments; x-scale and facets are appended by the caller.
line_plot <- function(data, y, ylab, x = n, ests = est_main, from_zero = TRUE) {
  ests <- intersect(names(lab_est), ests)                 # canonical order, no ghost keys
  p <- data |>
    filter(estimator %in% ests) |>
    ggplot(aes({{ x }}, {{ y }}, colour = estimator, group = estimator)) +
    geom_line(linewidth = 0.7) + geom_point(size = 1.8) +
    scale_colour_manual(values = pal_est, breaks = ests,
                        labels = \(b) unname(lab_est[b])) +
    labs(y = ylab, colour = NULL) + theme_sim
  if (from_zero) p + expand_limits(y = 0) else p
}

# ── 5. Twin panel: RMSE | |Bias|, one shared legend ──
panel <- function(data, oc, x = n, ests = est_main, extra = NULL,
                  title = oc_labs[oc], from_zero = TRUE, compact = FALSE) {
  d <- filter(data, outcome == oc)
  drop0 <- \(x) sub("^(-?)0\\.", "\\1.", x)
  yl <- if (compact) NULL else "RMSE"
  yr <- if (compact) NULL else "|Bias|"
  l <- line_plot(d, rmse,      yl, x = {{ x }}, ests = ests, from_zero = from_zero) + extra +
    theme(legend.position = "none")
  r <- line_plot(d, abs(bias), yr, x = {{ x }}, ests = ests, from_zero = from_zero) + extra +
    theme(legend.position = "none")
  if (compact) {
    l <- l + ggtitle("RMSE") + scale_y_continuous(labels = drop0)
    r <- r + ggtitle("|Bias|") + scale_y_continuous(labels = drop0)
  }
  (l | r) + plot_layout(guides = "collect") +
    plot_annotation(title = title) & theme(legend.position = "bottom")
}

# ── 6. MSE decomposition: Bias² / Variance share, stacked area ──
pal_comp <- c("Variance" = "#377eb8", "Bias\u00b2" = "#e41a1c")

decomp_plot <- function(data, ests = est_main, x = n, cols = outcome,
                        x_scale = scale_x_n, xlab = "n",
                        col_lab = labeller(outcome = oc_labs, n = lab_n),
                        title = NULL) {
  keep <- intersect(names(lab_est), ests)
  data |>
    filter(estimator %in% keep) |>
    mutate(denom    = bias^2 + empse^2,
           bias_sq  = bias^2  / denom,
           variance = empse^2 / denom,
           estimator = factor(lab_est[estimator], levels = lab_est[keep])) |>
    pivot_longer(c(variance, bias_sq), names_to = "component", values_to = "share") |>
    mutate(component = factor(component, c("variance","bias_sq"),
                              c("Variance","Bias\u00b2"))) |>
    ggplot(aes({{ x }}, share, fill = component)) +
    geom_area() +
    facet_grid(rows = vars(estimator), cols = vars({{ cols }}), switch = "y",
               labeller = col_lab) +
    x_scale +
    scale_y_continuous(breaks = c(1), labels = scales::percent) +
    scale_fill_manual(values = pal_comp) +
    labs(x = xlab, y = "Share of MSE", fill = NULL, title = title) +
    theme_sim +
    theme(strip.text.y.left = element_text(angle = 0))
}

# ── 7. Performance table: estimators × design columns, ordered by mean metric ──
# cols="n" → flat.  cols=c("overlap","n") → c-spanners.
# group="outcome" → stacks all outcomes in one table with row groups.
perf_table <- function(data, ests = est_main, cols = "n",
                       top = "rmse", bottom = "abs_bias",
                       digits = 2, order_by = "rmse",
                       lab = lab_est, n_lab_tbl = n_lab,
                       span_lab = \(x) paste0("c = ", x), span_desc = TRUE,
                       title = "", subtitle = NULL,
                       group = NULL, group_lab = oc_labs) {
  stopifnot(length(cols) %in% 1:2,
            c(top, bottom, order_by) %in% c("rmse","bias","abs_bias","empse"))
  keep <- intersect(names(lab), ests)
  cvar <- cols[length(cols)]
  svar <- if (length(cols) == 2L) cols[[1]] else NULL
  pm   <- c(rmse="RMSE", bias="Bias", abs_bias="|Bias|", empse="EmpSE")
  f    <- paste0("%.", digits, "f")
  grp  <- !is.null(group)
  
  d <- data |> filter(estimator %in% keep) |> mutate(abs_bias = abs(bias)) |> mutate(
    Estimator = factor(unname(lab[estimator]), levels = unname(lab[keep])),
    cell = sprintf(paste0(f, "<br><small>(", f, ")</small>"), .data[[top]], .data[[bottom]]),
    avg  = ave(.data[[order_by]], estimator, FUN = mean),
    .col = if (is.null(svar)) paste0(cvar, "=", .data[[cvar]])
    else paste0(svar, "=", .data[[svar]], "|", cvar, "=", .data[[cvar]]))
  if (grp) d <- mutate(d, Group = factor(group_lab[.data[[group]]], levels = group_lab))
  
  cv  <- sort(unique(d[[cvar]]))
  sv  <- if (is.null(svar)) NULL else sort(unique(d[[svar]]), decreasing = span_desc)
  key <- if (is.null(svar)) paste0(cvar, "=", cv)
  else c(t(outer(sv, cv, \(s, c) paste0(svar, "=", s, "|", cvar, "=", c))))
  
  id_vars <- c(if (grp) "Group", "Estimator", ".col", "cell", "avg")
  wide <- d |> distinct(across(all_of(id_vars))) |>
    pivot_wider(names_from = .col, values_from = cell)
  wide <- if (grp) arrange(wide, Group, avg) else arrange(wide, avg)
  present <- intersect(key, names(wide))
  wide <- wide |> select(any_of("Group"), Estimator, all_of(present))
  
  inner_lab <- \(x) { x <- sub("^[^=]*=", "", sub(".*\\|", "", x))
  if (is.null(n_lab_tbl)) x else unname(ifelse(x %in% names(n_lab_tbl), n_lab_tbl[x], x)) }
  
  g <- if (grp) gt(wide, groupname_col = "Group") else gt(wide)
  g <- g |> tab_header(title = title,
                       subtitle = md(subtitle %||% paste0(pm[[top]], " (", pm[[bottom]], ")")))
  for (s in sv) g <- tab_spanner(g, span_lab(s), starts_with(paste0(svar, "=", s, "|")))
  g |>
    cols_label_with(all_of(present), inner_lab) |>
    fmt_markdown(all_of(present)) |> cols_align("center", all_of(present)) |>
    tab_style(cell_text(weight = "bold"), cells_body(Estimator)) |>
    opt_table_lines("default") |>
    tab_source_note(md(sprintf("_Top %s / bottom %s; sorted by mean %s._",
                               pm[[top]], pm[[bottom]], pm[[order_by]])))
}


# ── 8. Convergence / path-truncation table ──
# % reps where ALL selected estimators coincide (range < tol).
# k=2: bal0 ≡ bal05.  k≥3: full-path collapse.  Raw replicates, not summaries.
converge_table <- function(data, ests, by = c("overlap","n"),
                           cols = by[length(by)], tol = 1e-9,
                           outcome_name = NULL, overlap_vals = NULL,
                           n_lab_tbl = n_lab, title = NULL) {
  stopifnot(length(ests) >= 2, length(cols) == 1, cols %in% by)
  d <- filter(data, estimator %in% ests,
              if (is.null(outcome_name)) TRUE else outcome == outcome_name,
              if (is.null(overlap_vals)) TRUE else overlap %in% overlap_vals)
  
  wide <- d |>
    group_by(across(all_of(c(by, "sim")))) |>
    summarise(hit = (max(tau_hat) - min(tau_hat)) < tol, .groups = "drop") |>
    group_by(across(all_of(by))) |>
    summarise(pct = sprintf("%.1f%%", mean(hit) * 100), .groups = "drop") |>
    pivot_wider(names_from = all_of(cols), values_from = pct, names_prefix = paste0(cols, "="))
  
  ord  <- grep(paste0("^", cols, "="), names(wide), value = TRUE)
  ord  <- ord[order(as.numeric(sub(".*=", "", ord)))]
  wide <- select(wide, all_of(setdiff(by, cols)), all_of(ord))
  if ("overlap" %in% names(wide)) names(wide)[names(wide) == "overlap"] <- "c"
  
  g <- gt(wide)
  if (!is.null(title)) g <- tab_header(g, title = md(title))
  g |>
    cols_label_with(all_of(ord), \(x) { v <- sub(".*=", "", x)
    if (is.null(n_lab_tbl)) v else unname(ifelse(v %in% names(n_lab_tbl), n_lab_tbl[v], v)) }) |>
    cols_align("center")
}

# ── 9. Misspecification helpers ──
# ΔRMSE and Δ|Bias| = misspecified − correct at matching (n, outcome, overlap, estimator).
delta_ms <- function(summ, ests = est_main) {
  summ |>
    filter(estimator %in% ests) |>
    mutate(absbias = abs(bias)) |>
    select(n, outcome, overlap, estimator, misspec, rmse, absbias) |>
    pivot_wider(names_from = misspec, values_from = c(rmse, absbias)) |>
    mutate(d_rmse = rmse_TRUE - rmse_FALSE,
           d_bias = absbias_TRUE - absbias_FALSE)
}

delta_rmse <- function(summ, ests = est_main)
  delta_ms(summ, ests) |> transmute(n, outcome, overlap, estimator, delta = d_rmse)

# §9 — compact ΔRMSE | Δ|Bias| twin, faceted by n, walked over outcomes
delta_panel <- function(data, oc, ests = est_main, ov = c(0.25, 0.5, 0.75, 1),
                        title = oc_labs[oc]) {
  drop0 <- \(x) sub("^(-?)0\\.", "\\1.", x)
  d <- filter(data, outcome == oc, overlap %in% ov)
  shared <- list(
    geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4),
    scale_x_reverse(breaks = ov),
    facet_wrap(~n, labeller = lab_n),
    scale_y_continuous(labels = drop0))
  l <- line_plot(d, d_rmse, NULL, x = overlap, ests = ests, from_zero = FALSE) +
    ggtitle("\u0394RMSE") + shared + theme(legend.position = "none")
  r <- line_plot(d, d_bias, NULL, x = overlap, ests = ests, from_zero = FALSE) +
    ggtitle("\u0394|Bias|") + shared + theme(legend.position = "none")
  (l | r) + plot_layout(guides = "collect") +
    plot_annotation(title = title) & theme(legend.position = "bottom")
}

# Style-(a): |Bias| and SD vs overlap, coloured by specification (Correct vs Misspecified).
spec_grid <- function(data, ests = est_bal, at_n = 5000, ov = c(0.25, 0.5, 0.75, 1, 2)) {
  ests <- intersect(names(lab_est), ests)
  data |>
    filter(estimator %in% ests, n == at_n, overlap %in% ov) |>
    pivot_longer(c(bias, empse), names_to = "metric") |>
    mutate(metric    = factor(metric, c("bias","empse"), c("|Bias|","SD")),
           spec      = ifelse(misspec, "Misspecified", "Correct"),
           estimator = factor(lab_est[estimator], levels = lab_est[ests]),
           outcome   = factor(outcome, names(oc_labs), oc_labs)) |>
    ggplot(aes(overlap, abs(value), colour = spec)) +
    geom_line() + geom_point(size = 1.5) +
    facet_grid(metric ~ estimator + outcome, scales = "free_y") +
    scale_x_continuous(breaks = c(0.25, 0.5, 1, 2), labels = c(".25", ".5", "1", "2")) +
    scale_colour_manual(values = pal_spec) +
    labs(x = "Overlap (c)", y = NULL, colour = NULL) +
    theme_bw(base_size = 10) +
    theme(legend.position = "bottom",
          strip.text.x = element_text(size = 7),
          axis.text.x  = element_text(angle = 45, hjust = 1, size = 7))
}


# ── 10. Extreme-estimate heatmap: reps with |tau_hat| > 10, overlap × n, faceted by outcome ──
# Tile colour + numerator = `est`'s extreme count per cell; denominator = `est`'s total across
# the whole figure. Labels only where the cell has ≥1 (empty cells left blank).
extreme_heat <- function(data, est = "glmcv_normal", ov = NULL,
                         high = "#B22222", title = NULL) {
  d <- data |>
    filter(estimator == est, if (is.null(ov)) TRUE else overlap %in% ov)
  total <- sum(d$n_extreme)                       # est's extreme estimates across the figure
  cells <- d |>
    group_by(outcome, n, overlap) |>
    summarise(extreme = sum(n_extreme), .groups = "drop") |>
    mutate(fill_val = if_else(extreme > 0, extreme, NA_real_),
           outcome  = factor(outcome, names(oc_labs), oc_labs),
           n        = factor(n, sort(unique(n))),
           overlap  = factor(overlap, sort(unique(overlap), decreasing = TRUE)))
  ggplot(cells, aes(overlap, n, fill = fill_val)) +
    geom_tile(colour = "white", linewidth = 0.6) +
    geom_text(aes(label = if_else(extreme > 0, sprintf("%.0f/%.0f", extreme, total), "")),
              size = 3) +
    facet_wrap(~outcome) +
    scale_fill_gradient(low = "white", high = high, name = "Extreme\n(count)", na.value = "white") +
    scale_y_discrete(labels = n_lab) +
    labs(x = "Overlap (c)", y = "n",
         title = title %||% sprintf("%s: extreme estimates (|\u03c4\u0302| > 10)",
                                    unname(lab_est[est]))) +
    theme_sim +
    theme(panel.grid = element_blank(), legend.position = "right")
}

# ============================================================================
# PART B — NEW SECTIONS (\u03bb paths, truncation, SMD/KKT, active set)
# ============================================================================

# ── 11. Diagnostics infrastructure ──
# Raw output is long (one row per rep x stat). diag_wide() lifts the
# diagnostic rows to one row per rep with named columns; as_diag() accepts
# either form so every helper below can be fed raw replicate data directly.
design_cols <- c("n","p","s","misspec","covcor","outcome",
                 "overlap","signs","strength")

diag_wide <- function(data) {
  id <- intersect(c(design_cols, "sim"), names(data))
  d  <- filter(data, estimator %in% diag_stats)
  if (!nrow(d)) stop("no diagnostic rows: file predates the logging DGP")
  pivot_wider(d, id_cols = all_of(id), names_from = estimator,
              values_from = tau_hat)
}

as_diag <- function(data)
  if (any(diag_stats %in% names(data))) data else diag_wide(data)

# Cell-level summaries: med / q10 / q90 of every logged quantity, plus
# convergence-failure count and truncation share.
summarise_diag <- function(data, by = design_cols) {
  dw <- as_diag(data)
  by <- intersect(by, names(dw))
  dw |>
    group_by(across(all_of(by))) |>
    summarise(
      m         = dplyr::n(),
      n_err     = sum(err == 1, na.rm = TRUE),
      p_trunc05 = mean(trunc05, na.rm = TRUE),
      across(matches("^(lam|nnz|smd)"),
             list(med = \(x) median(x, na.rm = TRUE),
                  q10 = \(x) quantile(x, .10, na.rm = TRUE, names = FALSE),
                  q90 = \(x) quantile(x, .90, na.rm = TRUE, names = FALSE))),
      .groups = "drop"
    )
}
# error screen:  summarise_diag(df) |> filter(n_err > 0)

# ── 12. \u03bb regularisation path / rate tracking ──
# Median (optional q10-q90 ribbon) of CV-selected \u03bb, attained path-endpoint \u03bb,
# and glmnet \u03bb.min vs n. Dashed black reference = \u221a(log p / n) (Wager 2024,
# section 7.2); dotted lines at the fixed-\u03bb marks. Colour = quantity,
# linetype = arm (CB loss fits one model per arm).
lam_meta <- tibble::tribble(
  ~stat,        ~quantity,           ~arm,
  "lam_balcv1", "CV \u03bb",         "treated",
  "lam_balcv0", "CV \u03bb",         "control",
  "lam_end1",   "path endpoint",     "treated",
  "lam_end0",   "path endpoint",     "control",
  "lam_glmcv",  "glmnet \u03bb.min", "single fit"
)
pal_lam <- c("CV \u03bb" = "#0072B2", "path endpoint" = "#009E73",
             "glmnet \u03bb.min" = "#B22222")
lt_arm  <- c(treated = "solid", control = "42", "single fit" = "solid")

lambda_plot <- function(dg, x = n, stats = lam_meta$stat,
                        ribbon = FALSE, rate = TRUE, ref_fixed = c(.05, .10)) {
  # rate reference assumes x = n; set rate = FALSE for any other x
  d <- dg |>
    pivot_longer(matches("^lam_[a-z0-9]+_(med|q10|q90)$"),
                 names_to = c("stat", ".value"),
                 names_pattern = "^(lam_[a-z0-9]+)_(med|q10|q90)$") |>
    filter(stat %in% stats) |>
    left_join(lam_meta, by = "stat")
  p <- ggplot(d, aes({{ x }}, med, colour = quantity, linetype = arm,
                     group = interaction(quantity, arm)))
  if (ribbon)
    p <- p + geom_ribbon(aes(ymin = q10, ymax = q90, fill = quantity,
                             group = interaction(quantity, arm)),
                         alpha = .12, colour = NA)
  if (rate) {
    rd <- dg |> distinct(across(any_of(design_cols))) |>
      mutate(med = sqrt(log(p) / n))
    p <- p + geom_line(data = rd, aes({{ x }}, med), inherit.aes = FALSE,
                       linetype = "dashed", colour = "black", linewidth = .5)
  }
  if (!is.null(ref_fixed))
    p <- p + geom_hline(yintercept = ref_fixed, linetype = "dotted",
                        colour = "grey45")
  p +
    geom_line(linewidth = .7) + geom_point(size = 1.5) +
    scale_colour_manual(values = pal_lam) +
    scale_fill_manual(values = pal_lam, guide = "none") +
    scale_linetype_manual(values = lt_arm) +
    scale_y_log10() +
    labs(y = "\u03bb", colour = NULL, linetype = NULL) + theme_sim
}
# dg <- summarise_diag(df)
# lambda_plot(dg) + scale_x_n +
#   facet_grid(overlap ~ outcome, labeller = labeller(outcome = oc_labs))
# CV-\u03bb level vs spread (slow decay vs CV-criterion instability):
# lambda_plot(dg, stats = c("lam_balcv1","lam_balcv0"), ribbon = TRUE) +
#   scale_x_n + facet_wrap(~outcome, labeller = as_labeller(oc_labs))

# ── 13. Path-truncation table (positivity diagnostic) ──
# % reps where the attained endpoint sits ABOVE \u03bb = 0.05 in either arm
# (trunc05 as logged). Direct replacement for inferring truncation from
# bal0 = bal05 estimate coincidence (converge_table, section 8, kept for
# old files).
trunc_table <- function(data, by = c("overlap","n"), cols = by[length(by)],
                        outcome_name = NULL, n_lab_tbl = n_lab, title = NULL) {
  dw <- as_diag(data)
  by <- intersect(by, names(dw))
  stopifnot(length(cols) == 1, cols %in% by)
  d <- dw |> filter(if (is.null(outcome_name)) TRUE else outcome == outcome_name)
  wide <- d |>
    group_by(across(all_of(by))) |>
    summarise(pct = sprintf("%.1f%%", 100 * mean(trunc05, na.rm = TRUE)),
              .groups = "drop") |>
    pivot_wider(names_from = all_of(cols), values_from = pct,
                names_prefix = paste0(cols, "="))
  ord <- grep(paste0("^", cols, "="), names(wide), value = TRUE)
  ord <- ord[order(as.numeric(sub(".*=", "", ord)))]
  wide <- select(wide, all_of(setdiff(by, cols)), all_of(ord))
  if ("overlap" %in% names(wide)) names(wide)[names(wide) == "overlap"] <- "c"
  g <- gt(wide)
  if (!is.null(title)) g <- tab_header(g, title = md(title))
  g |>
    cols_label_with(all_of(ord), \(x) { v <- sub(".*=", "", x)
    if (is.null(n_lab_tbl)) v else unname(ifelse(v %in% names(n_lab_tbl),
                                                 n_lab_tbl[v], v)) }) |>
    cols_align("center")
}
# trunc_table(df, by = c("overlap","n"), outcome_name = "linear",
#             title = "P(path endpoint \u03bb > 0.05)")

# ── 14. KKT / attained-imbalance check (balnet eq 10) ──
# Bound: attained max|SMD| <= \u03bb per arm. Reports the worst gap and the share
# of reps exceeding the bound by > tol. Gaps <= 0 with slack shrinking in n
# are the fixed-\u03bb plateau mechanism (imbalance converging up to the ceiling).
# NB the DGP logs a self-normalised SMD; small positive gaps may reflect that
# definitional mismatch rather than solver failure — confirm against the raw
# KKT form before flagging (open item noted in the run script).
kkt_check <- function(data, tol = 1e-6, by = design_cols) {
  dw <- as_diag(data)
  by <- intersect(by, names(dw))
  dw |>
    mutate(gap05 = pmax(smd1_05, smd0_05) - 0.05,
           gapcv = pmax(smd1_cv - lam_balcv1, smd0_cv - lam_balcv0)) |>
    group_by(across(all_of(by))) |>
    summarise(m           = dplyr::n(),
              max_gap_05  = max(gap05, na.rm = TRUE),
              pct_viol_05 = 100 * mean(gap05 > tol, na.rm = TRUE),
              max_gap_cv  = max(gapcv, na.rm = TRUE),
              pct_viol_cv = 100 * mean(gapcv > tol, na.rm = TRUE),
              .groups = "drop")
}

# Attained imbalance vs the fixed-\u03bb ceiling: med (q10-q90) of max|SMD| by
# arm; hline at .05 for at = "05". For at = "cv" the bound varies per rep —
# read jointly with kkt_check / lambda_plot.
smd_plot <- function(dg, x = n, at = "05", ribbon = TRUE) {
  d <- dg |>
    pivot_longer(matches(paste0("^smd[01]_", at, "_(med|q10|q90)$")),
                 names_to = c("stat", ".value"),
                 names_pattern = "^(smd[01]_[a-z0-9]+)_(med|q10|q90)$") |>
    mutate(arm = ifelse(startsWith(stat, "smd1"), "treated", "control"))
  yl <- if (at == "05") "attained max |SMD| at \u03bb = .05"
  else "attained max |SMD| at CV \u03bb"
  p <- ggplot(d, aes({{ x }}, med, linetype = arm, group = arm))
  if (ribbon) p <- p + geom_ribbon(aes(ymin = q10, ymax = q90, group = arm),
                                   alpha = .12, colour = NA)
  if (at == "05") p <- p + geom_hline(yintercept = .05, linetype = "dotted",
                                      colour = "grey45")
  p + geom_line(colour = "#0072B2", linewidth = .7) +
    geom_point(colour = "#0072B2", size = 1.5) +
    scale_linetype_manual(values = c(treated = "solid", control = "42")) +
    labs(y = yl, linetype = NULL) + theme_sim
}

# ── 15. Active-set size ──
lab_nnz <- c(nnz_balcv1 = "BalNet CV (treated)",
             nnz_balcv0 = "BalNet CV (control)",
             nnz_glm    = "glmnet \u03bb.min")
nnz_plot <- function(dg, x = n) {
  d <- dg |>
    pivot_longer(matches("^nnz_[a-z0-9]+_med$"),
                 names_to = "stat", values_to = "med") |>
    mutate(stat = sub("_med$", "", stat))
  ggplot(d, aes({{ x }}, med, colour = stat, group = stat)) +
    geom_line(linewidth = .7) + geom_point(size = 1.5) +
    scale_colour_manual(values = c(nnz_balcv1 = "#0072B2",
                                   nnz_balcv0 = "#56B4E9",
                                   nnz_glm    = "#B22222"),
                        labels = \(b) unname(lab_nnz[b])) +
    labs(y = "non-zero coefficients", colour = NULL) + the
```

## AR1: Correct Specification

```{r}
#| message: false
#| warning: false
# ── E1: Correctly specified DGP (strong overlap, c = 1) ──
df_e1 <- read_csv("ar1GDP1.csv.gz")

summ_e1 <- summarise_sim(df_e1) |> filter(!misspec, overlap == 1)
est_e1  <- c(est_main, "normalised_oracle")     # 5 main estimators + normalised oracle

# RMSE | |Bias| vs n, per outcome
walk(c("linear","quad1","exp"),
     \(oc) print(panel(summ_e1, oc, ests = est_e1, extra = scale_x_n)))


# RMSE (|Bias|) tables, per outcome
perf_table(summ_e1, ests = est_e1, group = "outcome", title = "E1: Correctly Specified")

# MSE decomposition (Bias²/Variance share), estimators × outcome (no oracle)
print(decomp_plot(summ_e1, ests = est_main,
                  title = "MSE Decomposition (%) \u2014 Correct spec (c = 1)"))
```

**Balancing vs MLE**

- **Under linear outcomes:** MLE and Set $\lambda \leq .05$ balancing estimators perform better under AR(1) correlation than when iid is assumed, both relatively and in absolute magnitude. Only Balnet (CV) is degraded ie:

  - at n=500 Bal0 estimated RMSE is 0.11, and 0.13 for iid and ar1 respecitvely. for bias, 0.01 vs 0.02.

  - at n=1000 normalized glmnet has RMSE 0.17 vs 0.2 with Bias 0.15 vs .18, ar1 and iid respectively.

  - CVbalnet: n=500, RMSE = 0.28 vs 0.24, Bias 0.25 vs 0.18, ar1 and iid respectively.

  - bal05 $\lambda_{endpoint}$ the same, convergence rate quicker. bal10 $\lambda_{endpoint}$ slightly higher.

  - by n=0,000)GLM (Hajek

  - gains are stronger for balancing estimators relative to MLE (bal05 outperforms MLE up to n=5000).

- **Under quadratic outcomes:** AR(1) correlation degrades, set Set $\lambda \leq .05$ balancing estimators perform better relative to MLE.

  - for bal0 and bal05 RMSE and bias absolute magnitude vs their iid equivilent are only moderately degraded, and could be sampling artifacts.

  - *for bal0, by n=50,000, it slightly outperforms iid \*pending MSCE decomp.*

  - MLE is degraded vs when iid at $n \leq 1000$

- exponential results are effectively the same as under linear outcomes.

## P\>N\>M

```{r}
# ── Dimensionality: correct spec, p/n sweep (c = 1) ──
dfm2 <- read_csv("HighDlowM.csv.gz")
summ_dim2 <- summarise_sim(dfm2) |> mutate(pn = round(p / n, 2))
dim_extra <- list(facet_wrap(~n, nrow = 1, labeller = lab_n, scales = "free_y"),
                  labs(x = "p/n"))
walk(c("linear", "quad1"), \(oc) print(
  panel(summ_dim2, oc, x = factor(pn), extra = dim_extra, from_zero = FALSE)))


perf_table(summ_dim2, cols = c("pn", "n"),
           span_lab = \(x) paste0("p/n = ", x), span_desc = FALSE,
           group = "outcome", title = "Dimensionality (c = 1)")

walk(c("linear", "quad1"), \(oc) print(
  decomp_plot(filter(summ_dim2, outcome == oc),
              x = pn, cols = n, xlab = "p/n",
              x_scale = scale_x_log10(breaks = c(0.5, 1, 2, 10, 20)),
              title = paste("MSE Decomposition (%) \u2014", oc_labs[oc]))))
```

Findings from n 500 to 1000 extended. minimum feasible $\lambda_{endpoint}$ associated with the balancing bias floor shrinks with $n$. by $p/n$ 20,

#### Between all

- **best estimators, bal0 and bal05,** perform identically across every p/n ration.

  - Bal10 is competitive at n=100. increases in n, decrease the $\lambda_{endpoint}$

- **Balnet. CV is the worst estimator.** MLE and bal.cv perform similarly under low n and high dimensionality. as sample size increases, MLE gains the edge .

- outcome complexity only has effect on magnitude and slope, not ranking.

## DGP2 CORRECT

```{r}
# ── Sign-pattern robustness (correct spec, c = 1, p = 50) ──
df_e1   <- read_csv("DGP2correct.csv.gz")


summ_e1 <- summarise_sim(df_e1) |> filter(!misspec, overlap == 1)
est_e1  <- c(est_main, "normalised_oracle")     # 5 main estimators + normalised oracle

# RMSE | |Bias| vs n, per outcome
walk(c("linear","quad1","exp"),
     \(oc) print(panel(summ_e1, oc, ests = est_e1, extra = scale_x_n)))

# RMSE (|Bias|) tables, per outcome
perf_table(summ_e1, ests = est_e1, group = "outcome", title = "E1: Correctly Specified")

# MSE decomposition (Bias²/Variance share), estimators × outcome (no oracle)
print(decomp_plot(summ_e1, ests = est_main,
                  title = "MSE Decomposition (%) \u2014 Correct spec (c = 1)"))


```

- setting a minimum for $bal.cv$ with `fit_cv <- cv.balnet(X, W, nfolds = 5, max.imbalance = 1e-4)` causes its relative performance to increase, with RMSE crossover happening at a earlier level.

  - Also confirms the $bal0$ bias flor was a tuning artifact, results are consistent with Wagner.

- changes in relative performance are due to overlap changing from sd-normalization.

## EFFECT OF SIGNS

```{r}
#overlap check
# e*(X) = {1 + exp(\u03b7 / (c\u00b7sd(\u03b7)))}^{-1},  \u03b7 = \u03a3_{j\u2264s} (\u03c3_j / j) X_j
sgn <- function(pat, s = 4)
  switch(pat, ks = c(1,-1,1,1)[1:s], pos = rep(1, s), mixed = rep(c(1,-1), length.out = s))

e_star <- function(X, pat, c = 1, s = 4) {
  eta <- drop(X[, 1:s, drop = FALSE] %*% (sgn(pat, s) / seq_len(s)))
  1 / (1 + exp(eta / (c * sd(eta))))
}

set.seed(1); N <- 1e5
X <- matrix(rnorm(N * 50), N, 50)          # iid, p = 50, as in the grid

bind_rows(lapply(names(lab_signs), \(sg) {
  e <- e_star(X, sg, c = 1)
  tibble(signs = lab_signs[[sg]],
         sd_index = sd(qlogis(1 - e)),                       # = 1/c exactly
         min_e = min(e), q01 = quantile(e, .01),
         q99 = quantile(e, .99), max_e = max(e),
         p_lt_10 = mean(e < .10), p_gt_90 = mean(e > .90))
}))

tibble(signs = rep(names(lab_signs), each = N),
       e     = unlist(lapply(names(lab_signs), \(sg) e_star(X, sg)))) |>
  mutate(signs = factor(lab_signs[signs], lab_signs)) |>
  ggplot(aes(e, colour = signs)) + geom_density(linewidth = .7) +
  labs(x = "true e*(X)", y = "density",
       title = "Overlap by sign pattern (iid, p = 50, c = 1)") + theme_sim
```

```{r}
# ── Sign-pattern robustness (correct spec, c = 1, iid, p = 50) ──
df_sg <- read_csv("DGP2covsign.csv.gz")                     
lab_signs <- c(ks = "K&S (+,\u2212,+,+)", pos = "All +", mixed = "Alternating")

summ_sg <- summarise_sim(df_sg) |>
  filter(signs %in% names(lab_signs)) |>
  mutate(signs = factor(signs, names(lab_signs), lab_signs))
est_sg <- c(est_main, "normalised_oracle")

# RMSE | |Bias| vs n, faceted by sign pattern, walked over outcomes
walk(c("linear","quad1"),
     \(oc) print(panel(summ_sg, oc, ests = est_sg,
                       extra = list(scale_x_n, facet_wrap(~signs)))))

# RMSE (|Bias|) table: outcome row-groups, sign spanners over n
perf_table(summ_sg |> filter(n %in% c(500, 5000,10000, 50000)),
           ests = est_sg, cols = c("signs","n"), group = "outcome",
           span_lab = identity, span_desc = FALSE,
           title = "Sign-pattern robustness: correct spec (c = 1)")

# MSE decomposition: estimators × sign pattern, per outcome
walk(c("linear","quad1"),
     \(oc) print(decomp_plot(filter(summ_sg, outcome == oc),
                             ests = est_main, cols = signs,
                             title = paste0("MSE Decomposition (%) \u2014 ",                                                               oc_labs[oc], " (c = 1)"))))


```

**Bal0's performance is invariant to sign changes,** confirming that exact balancing is insensitive to confounding alignment (the degree to which propensity coefficient signs match outcome coefficient signs).

**For our baseline-covariate sign profiles moving forward, we will use the fully aligned case,** $\bf \sigma = (1,1,....,1)$, where the sign of outcome and propensity score coefficients match perfectly. This is where the normalized MLE's relative performance over balancing is strongest (exception at small $n$, but doesnt change observed crossovers).

- MLE's relative advantage increases with $n$, and at low $n$ its performance gaps are actually the most pronounced.

**balnet.cv** as alignment *increases*, balnetcv improves relative to bal05/bal10 (their bias explodes faster than its own)

- no reason to worry about sign effects again, summarise this all better in the appendix.

- prior results with DGP 1 use (K&S from tan paper).

`DGP2corsignTEST` shows thats under AR(1) correlated covariates these sign can effects generally hold (interaction effects change sign)

## SPARSITY AND SIGNAL 

NOTE: d
